---
title: "Bioequivalence Tests for Parallel Trial Designs with Log-Normal Data"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bioequivalence Tests for Parallel Trial Designs with Log-Normal Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: 'references.bib'
link-citations: yes
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
options(rmarkdown.html_vignette.check_title = FALSE) #title of doc does not match vignette title
doc.cache <- T #for cran; change to F
```

In the `SimTOST` R package, which is specifically designed for sample size estimation for bioequivalence studies, hypothesis testing is based on the Two One-Sided Tests (TOST) procedure. [@sozu_sample_2015]  In TOST, the equivalence test is framed as a comparison between the the null hypothesis of ‘new product is worse by a clinically relevant quantity’ and the alternative hypothesis of ‘difference between products is too small to be clinically relevant’. This vignette focuses on a parallel design, with 2 arms/treatments and 5 primary endpoints.

In the following two examples, we demonstrate the use of **SimTOST** for parallel trial designs with data assumed to follow a normal distribution on the log scale. We start by loading the package.

The examples follow a practical simulation workflow. We first specify the
estimand, margins, outcome distribution, and endpoint dependence; next we
estimate sample size or achieved power; finally we inspect the retained
simulations. The diagnostics are deliberately separated from the power result:
they help determine whether the simulated data match the assumptions used to
obtain that result.

```{r, echo = T, message=F}
library(SimTOST)
```

# Multiple Independent Co-Primary Endpoints
Here, we consider a bio-equivalence trial with 2 treatment arms and $m=5$ endpoints. The sample size is calculated to ensure that the test and reference products are equivalent with respect to all 5 endpoints. The true ratio between the test and reference products is assumed to be 1.05. Outcomes are simulated on the original scale from a log-normal distribution, with a standard deviation of $0.3$ on the log scale and independent endpoints ($\rho = 0$). The equivalence limits are set at 0.80 and 1.25. The significance level is 0.05. The sample size is determined at a power of 0.8.

This example is adapted from @mielke_sample_2018, who employed a difference-of-means test on the log scale. The log-normal analysis below uses a ratio-of-means TOST, which is equivalent to testing the difference on the log scale under the log-normal model.

## Approach 1: Using sampleSize_Mielke
In the first approach, we calculate the required sample size for 80% power using the [sampleSize_Mielke()](../reference/sampleSize_Mielke.html) function. This method directly follows the approach described in @mielke_sample_2018, assuming a difference-of-means test on the log-transformed scale with specified parameters.

```{r, eval = TRUE}
ssMielke <- sampleSize_Mielke(power = 0.8, Nmax = 1000, m = 5, k = 5, rho = 0,
                              sigma = 0.3, true.diff = log(1.05),
                              equi.tol = log(1.25), design = "parallel",
                              alpha = 0.05, adjust = "no", seed = 1234,
                              nsim = 1000)
ssMielke
```
```{r mielke-value, include = FALSE}
ss_value <- ssMielke$SS
```
For 80\% power, `r ss_value` subjects per arm (`r ss_value * 2` in total) would be required.

## Approach 2: Using sampleSize
Alternatively, the sample size calculation can be performed using the [sampleSize()](../reference/sampleSize.html) function. This method simulates log-normal outcomes on the original scale and applies a ratio-of-means test (`ctype = "ROM"`). The means and standard deviations below are supplied on the original scale; a log-scale standard deviation of 0.3 corresponds to the displayed coefficient of variation.

```{r, eval = TRUE}
mu_r <- setNames(rep(1.00, 5), paste0("y", 1:5))
mu_t <- setNames(rep(1.05, 5), paste0("y", 1:5))
sigma <- setNames(sqrt(exp(0.3^2) - 1) * mu_r, paste0("y", 1:5))
lequi_lower <- setNames(rep(0.8, 5), paste0("y", 1:5))
lequi_upper <- setNames(rep(1.25, 5), paste0("y", 1:5))

ss <- sampleSize(power = 0.8, alpha = 0.05,
                 mu_list = list("R" = mu_r, "T" = mu_t),
                 sigma_list = list("R" = sigma, "T" = sigma),
                 list_comparator = list("R_vs_T" = c("R", "T")),
                 list_lequi.tol = list("R_vs_T" = lequi_lower),
                 list_uequi.tol = list("R_vs_T" = lequi_upper),
                 dtype = "parallel", ctype = "ROM", distribution = "lnorm",
                 adjust = "no", ncores = 1, nsim = 1000, seed = 1234,
                 keep_sim_data = TRUE)
ss
```

For a two-arm parallel design, the reported sample size of sampleSize() function is the total number of subjects (`r ss$response$n_total`), which corresponds to `r ss$response$n_total/2` subjects per arm. This is close to the number of patients obtained using the Mielke calculation.
The two results need not be identical: `sampleSize_Mielke()` simulates a
known-variance normal test
statistic, while `sampleSize()` simulates raw normal observations, estimates
the variance in each trial, and uses a t-based TOST. Monte Carlo error also
affects both estimates.


## Estimating power at a fixed sample size

In scenarios with economic or ethical constraints, the sample size may be fixed. In such cases, you can evaluate the achieved power for the current configuration using `simPower()`.
For example, you can evaluate multiple base sample sizes (n = 50, 100) per arm using the same log-normal assumptions. The resulting plot displays total sample size on the x-axis and corresponding achieved power in y-axis.

```{r fixed-power-parallel, eval = TRUE}
fixed_power <- simPower(
  n = c(50,100),
  distribution = "lnorm",
  mu_list = list(R = mu_r, T = mu_t),
  sigma_list = list(R = sigma, T = sigma),
  list_comparator = list(R_vs_T = c("R", "T")),
  list_y_comparator = list(R_vs_T = paste0("y", 1:5)),
  list_lequi.tol = list(R_vs_T = lequi_lower),
  list_uequi.tol = list(R_vs_T = lequi_upper),
  dtype = "parallel", ctype = "ROM", nsim = 1000, seed = 1234,
  keep_sim_data = TRUE
)
fixed_power
```

The returned object contains the estimated power and its Monte Carlo
confidence interval in `fixed_power$power`, `fixed_power$power_LCI`, and
`fixed_power$power_UCI`. Because `keep_sim_data = TRUE`, it also contains the
retained model-scale observations in `fixed_power$sim_data`. Retaining these
observations increases memory use and is therefore disabled by default.

## Methods for inspecting the sample-size result

The `sampleSize()` result is an object of class `simss`. The standard methods
provide complementary views of the same simulation:

```{r unified-result-methods, eval = TRUE, fig.width = 8, fig.height = 4.8,
      out.width = "100%", fig.align = "center"}
print(ss)
summary(ss)
confint(ss)
plot(ss)
```

`print(ss)` gives a compact overview of the design, estimand, margins, sample
size, and achieved power. `summary(ss)` returns the main design and power
information in a structured form. `confint(ss)` extracts the Monte Carlo
confidence interval for achieved power. `plot(ss)` displays the achieved
power and its confidence interval across the candidate sample sizes; for a
parallel design, the x-axis uses total sample size. These intervals quantify
simulation uncertainty and are not confidence intervals for the treatment
effect.

# Multiple Correlated Co-Primary Endpoints
In the second example, we set $k=m=5$, $\sigma = 0.3$ and $\rho = 0.8$. This example is also adapted from @mielke_sample_2018, who employed a difference-of-means test on the log scale. The sample size calculation can again be conducted using two approaches, both of which are illustrated below.

## Approach 1: Using sampleSize_Mielke
In the first approach, we calculate the required sample size for 80% power using the [sampleSize_Mielke()](../reference/sampleSize_Mielke.html) function. This method directly follows the approach described in @mielke_sample_2018, assuming a difference-of-means test on the log-transformed scale with specified parameters.

```{r, eval = TRUE}
ssMielke <- sampleSize_Mielke(power = 0.8, Nmax = 1000, m = 5, k = 5, rho = 0.8,
                              sigma = 0.3, true.diff = log(1.05),
                              equi.tol = log(1.25), design = "parallel",
                              alpha = 0.05, adjust = "no", seed = 1234,
                              nsim = 10000)
ssMielke
ss_value <- if (is.list(ssMielke)) ssMielke$SS else unname(ssMielke["SS"])
```
For 80\% power, `r ss_value` subjects per arm (`r ss_value * 2` in total) would be required.

## Approach 2: Using sampleSize
Alternatively, the sample size calculation can be performed using the [sampleSize()](../reference/sampleSize.html) function. This method simulates log-normal outcomes on the original scale and uses a ratio-of-means test (`ctype = "ROM"`) with user-specified values for `mu_list`, `sigma_list`, and the correlation parameter `rho`.

```{r}
mu_r <- setNames(rep(1.00, 5), paste0("y", 1:5))
mu_t <- setNames(rep(1.05, 5), paste0("y", 1:5))
sigma <- setNames(sqrt(exp(0.3^2) - 1) * mu_r, paste0("y", 1:5))
lequi_lower <- setNames(rep(0.8, 5), paste0("y", 1:5))
lequi_upper <- setNames(rep(1.25, 5), paste0("y", 1:5))

ss <- sampleSize(power = 0.8, alpha = 0.05,
                 mu_list = list("R" = mu_r, "T" = mu_t),
                 sigma_list = list("R" = sigma, "T" = sigma),
                 rho = 0.8, # high correlation between the endpoints
                 list_comparator = list("R_vs_T" = c("R", "T")),
                 list_lequi.tol = list("R_vs_T" = lequi_lower),
                 list_uequi.tol = list("R_vs_T" = lequi_upper),
                 dtype = "parallel", ctype = "ROM", distribution = "lnorm",
                 adjust = "no", ncores = 1, k = 5, nsim = 10000, seed = 1234)
ss
```

To verify the non-zero dependence assumption in this second scenario, retain a
moderate number of simulated trials at a fixed sample size. The correlation
plot is a model diagnostic: it shows whether the observed endpoint
correlations fluctuate around the value implied by `rho = 0.8`.

```{r correlated-endpoint-correlation, eval = TRUE, fig.width = 8, fig.height = 5, out.width = "100%", fig.align = "center"}
correlated_power <- simPower(
  n = 100,
  distribution = "lnorm",
  mu_list = list(R = mu_r, T = mu_t),
  sigma_list = list(R = sigma, T = sigma),
  rho = 0.8,
  list_comparator = list(R_vs_T = c("R", "T")),
  list_lequi.tol = list(R_vs_T = lequi_lower),
  list_uequi.tol = list(R_vs_T = lequi_upper),
  dtype = "parallel", ctype = "ROM", nsim = 500, seed = 1234,
  keep_sim_data = TRUE
)
plot_distribution(correlated_power, estimand = "correlation",
                  arms = c("R", "T"))
```

The reference line is the correlation of the simulated endpoints, not the
correlation between treatment arms. Arms are simulated independently; the
dependence being checked is between endpoints measured on the same participant
within an arm.

# References
