---
title: "Count-Outcome Equivalence: Parallel Design with 3 Arms and 3 Endpoints"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Count-Outcome Equivalence: Parallel Design with 3 Arms and 3 Endpoints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      warning = FALSE, message = FALSE)
library(SimTOST)
```

## Overview

This example illustrates how to plan a parallel count-outcome equivalence
study with three treatment arms and three count endpoints. The arms are a test
product (`TEST`) and two reference products (`EU_REF` and `US_REF`). We use
three clinically interpretable, illustrative endpoint names: `Exacerbations`,
`Hospitalizations`, and `RescueEvents` (rescue-medication events). These labels
represent event counts collected over the same follow-up period. The general
distributional assumptions used by all
SimTOST outcomes are described in the companion vignette
[`methodological_assumptions.Rmd`](methodological_assumptions.Rmd).

The count module supports scalar or vector-valued endpoint inputs. A joint
simulation can require equivalence for `k` of the `m` endpoints and can apply
Bonferroni, Mielke's strong k-out-of-m (`adjust = "t"`), or Šidák adjustment
to the endpoint-wise one-sided significance level. For a `k`-of-`m` decision,
the t-adjustment uses `alpha / (m - k + 1)` [@mielke_sample_2018]. The joint
count kernel can also simulate correlated endpoints through
the `cor_mat` argument.

## Study assumptions

The assumed event rates are:

```{r rates}
rate_list <- list(
  TEST = c(Exacerbations = 0.19,
           Hospitalizations = 0.13,
           RescueEvents = 0.08),
  EU_REF = c(Exacerbations = 0.20,
             Hospitalizations = 0.13,
             RescueEvents = 0.08),
  US_REF = c(Exacerbations = 0.21,
             Hospitalizations = 0.12,
             RescueEvents = 0.08)
)

comparisons <- list(
  EU_comparison = c("TEST", "EU_REF"),
  US_comparison = c("TEST", "US_REF")
)

lower_margin <- 0.80
upper_margin <- 1.25
# Five units of follow-up provide a realistic event-count scale for this
# illustrative example and allow the target power to be reached efficiently.
exposure <- 5
```

The rate-ratio equivalence interval is 0.80 to 1.25. We use a Poisson model
for the primary example and a one-sided significance level of 0.05 for each
TOST component.

## Power for a fixed sample size

Before estimating sample size, power can be examined for each endpoint and
comparison. Here we use 100 participants per treatment arm and a small number
of simulations to keep the vignette fast.

```{r fixed-power}
fixed_power <- do.call(rbind, lapply(names(comparisons), function(comp) {
  arms <- comparisons[[comp]]
  do.call(rbind, lapply(names(rate_list[[arms[1]]]), function(endpoint) {
    result <- simPower(
      n = 100,
      distribution = "pois",
      rate_list = setNames(list(
        setNames(rate_list[[arms[1]]][[endpoint]], endpoint),
        setNames(rate_list[[arms[2]]][[endpoint]], endpoint)
      ), arms),
      list_comparator = setNames(list(arms), comp),
      list_lequi.tol = setNames(list(lower_margin), comp),
      list_uequi.tol = setNames(list(upper_margin), comp),
      exposure = exposure,
      dtype = "parallel",
      nsim = 1000,
      seed = 1234
    )
    data.frame(comparison = comp, endpoint = endpoint,
               power = result$power,
               power_LCI = result$power_LCI,
               power_UCI = result$power_UCI)
  }))
}))

fixed_power
```

## Sample-size estimation for all comparisons and endpoints

We now estimate the smallest sample size per arm that reaches 80% power for
each individual comparison and endpoint.

```{r sample-size}
sample_size_results <- do.call(rbind, lapply(names(comparisons), function(comp) {
  arms <- comparisons[[comp]]
  do.call(rbind, lapply(names(rate_list[[arms[1]]]), function(endpoint) {
    result <- sampleSize(
      power = 0.80,
      distribution = "pois",
      rate_list = setNames(list(
        setNames(rate_list[[arms[1]]][[endpoint]], endpoint),
        setNames(rate_list[[arms[2]]][[endpoint]], endpoint)
      ), arms),
      list_comparator = setNames(list(arms), comp),
      list_lequi.tol = setNames(list(lower_margin), comp),
      list_uequi.tol = setNames(list(upper_margin), comp),
      exposure = exposure,
      dtype = "parallel",
      nsim = 1000,
      seed = 1234,
      lower = 10,
      upper = 2000
    )
    data.frame(comparison = comp, endpoint = endpoint,
               n_per_arm = result$n_per_arm,
               n_total_for_pair = result$n_total,
               achieved_power = result$power)
  }))
}))

sample_size_results

stopifnot(all(sample_size_results$achieved_power >= 0.80))
```

For a simple conservative planning rule, select the maximum required sample
size per arm across all six comparison-endpoint combinations:

```{r conservative-size}
required_per_arm <- max(sample_size_results$n_per_arm)
required_total <- 3 * required_per_arm

c(required_per_arm = required_per_arm, required_total = required_total)
```

The total is multiplied by three because the study has three treatment arms.
This rule ensures that each individual comparison and endpoint has at least
the target simulated power under its own assumptions. It is not a substitute
for a multiplicity-adjusted joint power calculation.

## Joint sample-size estimation with `k = 3`

The joint sample-size function receives all three arms, both comparison
families, and all three endpoints at once. Here, `k = 3` requires all three
endpoints to demonstrate equivalence for every comparison, while the
Bonferroni option adjusts across all six comparison-endpoint tests. The
returned sample size is therefore based on one joint simulated success
criterion rather than the maximum of separate searches.

```{r joint-sample-size}
endpoint_corr <- matrix(c(
  1.0, 0.40, 0.25,
  0.40, 1.0, 0.35,
  0.25, 0.35, 1.0
), nrow = 3, byrow = TRUE)

joint_result <- sampleSize(
  power = 0.80,
  distribution = "pois",
  rate_list = rate_list,
  list_comparator = comparisons,
  list_lequi.tol = list(
    EU_comparison = rep(lower_margin, 3),
    US_comparison = rep(lower_margin, 3)
  ),
  list_uequi.tol = list(
    EU_comparison = rep(upper_margin, 3),
    US_comparison = rep(upper_margin, 3)
  ),
  exposure = rep(exposure, 3),
  cor_mat = endpoint_corr,
  dtype = "parallel",
  nsim = 500,
  seed = 1234,
  lower = 10,
  upper = 3000,
  k = 3,
  adjust = "bonferroni"
)

joint_sample_size <- data.frame(
  n_per_arm = joint_result$n_per_arm,
  n_total = joint_result$n_total,
  achieved_power = joint_result$power,
  k = joint_result$k,
  adjustment = joint_result$adjust
)

joint_sample_size
stopifnot(all(joint_sample_size$achieved_power >= 0.80))
```

For a three-arm allocation, the reported total is three times the selected
number per arm. Both comparison families share the simulated test-arm counts,
and the endpoint correlation is generated through a Gaussian-copula count
model. This is therefore a joint count simulation rather than a maximum of
separate comparison-specific searches.

## Why joint planning is important

The separate calculation above targets 80% power for each comparison-endpoint
combination. That does not mean that the complete trial has 80% probability of
passing all six requirements. For example, if two requirements each have 80%
power and are independent, their joint success probability is only
$0.80 \times 0.80 = 0.64$.

The following comparison evaluates the separate result under the actual joint
criterion. The joint calculations use Bonferroni adjustment across the six
comparison-endpoint tests and require all three endpoints for both comparison
families (`k = 3`). The independent-endpoint scenario uses an identity
correlation matrix; the correlated scenario uses the matrix specified above.

```{r joint-advantage}
joint_power_at_separate <- simPower(
  n = required_per_arm,
  distribution = "pois",
  rate_list = rate_list,
  list_comparator = comparisons,
  list_lequi.tol = list(
    EU_comparison = rep(lower_margin, 3),
    US_comparison = rep(lower_margin, 3)
  ),
  list_uequi.tol = list(
    EU_comparison = rep(upper_margin, 3),
    US_comparison = rep(upper_margin, 3)
  ),
  exposure = rep(exposure, 3),
  cor_mat = endpoint_corr,
  dtype = "parallel",
  nsim = 1000,
  seed = 1234,
  k = 3,
  adjust = "bonferroni"
)

joint_independent_result <- sampleSize(
  power = 0.80,
  distribution = "pois",
  rate_list = rate_list,
  list_comparator = comparisons,
  list_lequi.tol = list(
    EU_comparison = rep(lower_margin, 3),
    US_comparison = rep(lower_margin, 3)
  ),
  list_uequi.tol = list(
    EU_comparison = rep(upper_margin, 3),
    US_comparison = rep(upper_margin, 3)
  ),
  exposure = rep(exposure, 3),
  cor_mat = diag(3),
  dtype = "parallel",
  nsim = 500,
  seed = 1234,
  lower = 10,
  upper = 3000,
  k = 3,
  adjust = "bonferroni"
)

joint_comparison <- data.frame(
  approach = c(
    "Separate searches; joint power evaluated afterward",
    "Joint search; independent endpoints",
    "Joint search; correlated endpoints"
  ),
  n_per_arm = c(
    required_per_arm,
    joint_independent_result$n_per_arm,
    joint_result$n_per_arm
  ),
  joint_power = c(
    joint_power_at_separate$power,
    joint_independent_result$power,
    joint_result$power
  )
)

joint_comparison
```

The separate-search sample size is smaller because it guarantees only the
individual 80% targets. In this example, its joint power is approximately 5%
after the six-test Bonferroni adjustment, whereas the joint searches achieve
the requested 80% trial-level power. This is the main advantage of joint
planning: it answers the question that matters for the confirmatory trial,
namely the probability that all required comparisons and endpoints succeed in
the same simulated study. Positive endpoint correlation can reduce the joint
sample size because endpoint successes tend to occur together, but it does not
remove the need for a joint calculation.

## Negative-binomial sensitivity analysis

If prior evidence suggests overdispersion, repeat the calculations with the
negative-binomial model. The `dispersion` parameter controls the amount of
overdispersion; larger values imply greater variability.

```{r negative-binomial-sensitivity}
nb_sensitivity <- simPower(
  n = required_per_arm,
  distribution = "nbinom",
  rate_list = list(
    TEST = rate_list$TEST[["Exacerbations"]],
    EU_REF = rate_list$EU_REF[["Exacerbations"]]
  ),
  list_comparator = list(TEST_vs_EU = c("TEST", "EU_REF")),
  list_lequi.tol = list(TEST_vs_EU = lower_margin),
  list_uequi.tol = list(TEST_vs_EU = upper_margin),
  exposure = exposure,
  dtype = "parallel",
  dispersion = 0.10,
  nsim = 300,
  seed = 1234
)

nb_sensitivity
```

## Interpretation and limitations

This vignette demonstrates both a conservative and a joint endpoint workflow
for a three-arm, three-endpoint count study using the current API. In the
separate-search workflow, endpoint and comparison calculations are performed
independently, whereas the joint workflow evaluates the required success event
in the same simulated trial. The distributional and dependence assumptions
are described in the companion
[`methodological_assumptions.Rmd`](methodological_assumptions.Rmd) vignette.

For a confirmatory three-arm/three-endpoint trial, the statistical analysis
plan should specify the familywise error strategy and joint success criterion.
