---
title: "Count Outcomes"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: 'references.bib'
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Count Outcomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE,
                      message = FALSE)
library(SimTOST)
```

## Overview

This vignette follows the workflow in `workflow.Rmd` for count outcomes. It
starts with a published two-arm Poisson benchmark, then validates a
two-endpoint Poisson model, calculates its required sample size, repeats the
calculation under a negative-binomial model, and finally uses the same model
settings for a balanced two-by-two crossover design.

For count outcomes, the estimand is the event-rate ratio

\[
RR = \frac{\lambda_T}{\lambda_R},
\]

where the first arm in each comparator is the test arm and the second is the
reference arm. Equivalence is assessed using two one-sided tests on the
log-rate-ratio scale. The workflow separates model validation from the final
equivalence calculation: first check the supplied parameters and retained
simulated data, then calculate and interpret the sample size.

## A Published Poisson benchmark

Zhu [@zhu_sample_2017] reports a two-arm Poisson equivalence calculation with
equal event rates of 1 per time unit, average exposure of 0.7, equivalence
limits of 0.9 and $1/0.9$, one-sided $\alpha=0.025$, and 2,705 participants
per arm. The reported total sample size is 5,410, with approximated power
0.80012.

```{r poisson-literature-benchmark}
zhu_benchmark <- simPower(
  n = 2705,
  distribution = "pois",
  rate_list = list(TEST = 1, REF = 1),
  list_comparator = list(TEST_vs_REF = c("TEST", "REF")),
  list_lequi.tol = list(TEST_vs_REF = 0.9),
  list_uequi.tol = list(TEST_vs_REF = 1 / 0.9),
  exposure = 0.7,
  dtype = "parallel",
  alpha = 0.025,
  nsim = 5000,
  seed = 2024
)
zhu_benchmark
data.frame(reference_power = 0.80012,
           simulated_power = zhu_benchmark$power,
           simulated_lower = zhu_benchmark$power_LCI,
           simulated_upper = zhu_benchmark$power_UCI)
```

The simulation need not reproduce the published value exactly. The published
result is a large-sample approximation, whereas `simPower()` simulates
discrete event totals and reports a finite-sample Monte Carlo estimate.

## A Poisson study
### Step 1: Define the study and theoretical targets

We now use the two-endpoint settings used throughout the rest of this
vignette. The supplied endpoint correlation is a latent Gaussian-copula
correlation, not the expected Pearson correlation of the observed counts.

```{r poisson-inputs}
count_corr <- matrix(c(1, 0.5, 0.5, 1), nrow = 2,
                     dimnames = list(c("y1", "y2"), c("y1", "y2")))
count_rates <- list(TEST = c(y1 = 0.21, y2 = 0.24),
                    REF = c(y1 = 0.20, y2 = 0.22))
count_comparators <- list(TEST_vs_REF = c("TEST", "REF"))
count_lower <- list(TEST_vs_REF = c(y1 = 0.80, y2 = 0.80))
count_upper <- list(TEST_vs_REF = c(y1 = 1.25, y2 = 1.25))
count_exposure <- 10
```

### Step 2: Run a pilot with sampleSize()
As in the workflow vignette, first run a fixed-sample pilot while retaining
the simulated data. This object is used for the diagnostics below.

```{r poisson-diagnostics}
poisson_diagnostic <- sampleSize(
  distribution = "pois", rate_list = count_rates,
  list_comparator = count_comparators,
  list_lequi.tol = count_lower, list_uequi.tol = count_upper,
  exposure = count_exposure, cor_mat = count_corr,
  dtype = "parallel", nsim = 500, seed = 1234,
  keep_sim_data = TRUE
)
poisson_diagnostic
confint(poisson_diagnostic)
```


### Step 3: Validate the generated data and estimands

The workflow checks result precision, marginal outcomes, exposure-adjusted
rates, endpoint dependence, and Monte Carlo stability before planning.

#### 3a. Arm-specific parameters


```{r poisson-rate-distribution, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(poisson_diagnostic, estimand = "rate")
```

The empirical rate distributions should be centered near the values in
`count_rates`. 

#### 3b. Endpoint correlations
```{r poisson-correlation-distribution, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(poisson_diagnostic, estimand = "correlation",
                  arms = c("TEST", "REF"))
```
The dashed line is the latent correlation supplied through `count_corr`. The
observed Pearson correlation can differ because the latent normal variables
are transformed into discrete counts. The plot checks the direction and
approximate magnitude of the generated dependence; `endpoint_corr` is the
exact parameter check.

#### 3c. Estimand direction and sampling distribution

```{r poisson-outcome-distribution, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(poisson_diagnostic, estimand = "outcome", type = "histogram")
```
The empirical counts should be compatible with the Poisson reference
distribution.

```{r poisson-estimand-distribution, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(poisson_diagnostic, estimand = "RR")
```


### Step 4: Check Monte Carlo stability and precision

```{r poisson-stability, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_stability(poisson_diagnostic)
plot_mc_error(poisson_diagnostic)
```


### Step 5: Check Type I error at the equivalence boundaries
```{r poisson-type1, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
type1_one <- type1Error(x = poisson_diagnostic, null = "both", joint = TRUE)
plot(type1_one)
```


### Step 6: Run the final sample-size calculation
This call uses exactly the rates, exposure, correlation matrix, margins, and
parallel design used by `poisson_diagnostic`.

```{r poisson-sample-size}
poisson_sample_size <- sampleSize(
  power = 0.80, distribution = "pois", rate_list = count_rates,
  list_comparator = count_comparators,
  list_lequi.tol = count_lower, list_uequi.tol = count_upper,
  exposure = count_exposure, cor_mat = count_corr, dtype = "parallel",
  lower = 100, upper = 1000, nsim = 1000, seed = 1234, ncores = 1,
  keep_sim_data = TRUE
)
summary(poisson_sample_size)
confint(poisson_sample_size)
```

For final planning, increase `nsim` and independently verify the selected
sample size with `simPower()`.


```{r poisson-result, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot(poisson_sample_size)
```


### Repeat with a negative-binomial distribution

The negative-binomial calculation keeps the same rates, exposure, correlation,
margins, and parallel design. The dispersion parameter allows the count
variance to exceed its Poisson value. We use `dispersion = 0.50` here as a
deliberately visible sensitivity example. For a mean count near 2, the
Poisson variance is about 2, whereas the negative-binomial variance is
approximately `2 + 0.50 * 2^2 = 4`.

```{r negative-binomial-sample-size}
nb_dispersion <- 0.50
negative_binomial_sample_size <- update(poisson_sample_size,
                                        distribution = "nbinom",
                                        dispersion = nb_dispersion)

summary(negative_binomial_sample_size)
confint(negative_binomial_sample_size)
```

```{r nb-outcome-distribution, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(negative_binomial_sample_size, estimand = "outcome", type = "histogram")
```

The result is conditional on `nb_dispersion`; vary it in sensitivity analyses
when overdispersion is uncertain.

### Use the same settings for a two-by-two crossover

Finally, we retain the negative-binomial rates, exposure, dispersion,
correlation matrix, margins, and target power, and change only the design to a
balanced two-by-two crossover. For `dtype = "2x2"`, `n` and `n_per_arm` refer
to participants per sequence.

```{r crossover-negative-binomial-sample-size}

crossover_sample_size <- update(negative_binomial_sample_size,
                                dtype = "2x2")

summary(crossover_sample_size)
confint(crossover_sample_size)
```

The crossover calculation uses the same treatment-effect and endpoint
settings but evaluates paired treatment-period data. Additional assumptions
can be supplied through `sigmaB`, `Eper`, `Eco`, and `dropout` when supported
by the study design or prior data.

## Interpretation and scope

The Poisson, negative-binomial, and crossover results are conditional on the
specified rates, exposure, margins, endpoint dependence, dispersion, and
design. Compare sample sizes only after the corresponding model checks pass.
For count outcomes, `cor_mat` is a latent Gaussian-copula correlation matrix;
it is not a correlation between treatment arms and is not required to equal
the Pearson correlation of the observed counts.
