---
title: "Count Outcomes"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: 'references.bib'
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Count Outcomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE,
                      message = FALSE)
library(SimTOST)
```

## Overview

This vignette follows the workflow in `workflow.Rmd` for count outcomes. It
starts with a published two-arm Poisson benchmark, then validates a
two-endpoint Poisson model, calculates its required sample size, repeats the
calculation under a negative-binomial model, and finally uses the same model
settings for a balanced two-by-two crossover design.

For count outcomes, the estimand is the event-rate ratio

\[
RR = \frac{\lambda_T}{\lambda_R},
\]

where the first arm in each comparator is the test arm and the second is the
reference arm. Equivalence is assessed using two one-sided tests on the
log-rate-ratio scale. The workflow separates model validation from the final
equivalence calculation: first check the supplied parameters and retained
simulated data, then calculate and interpret the sample size.

### Parameters required for count outcomes

Count models require an event rate for every arm and endpoint, rather than a
mean and standard deviation as in a continuous-outcome model. The main inputs
are:

* `rate_list`: the event rate for each arm, expressed per unit of exposure;
* `exposure`: the amount of observation time or opportunity contributed by
  one participant;
* `list_comparator`: the arm pairs to compare;
* `list_lequi.tol` and `list_uequi.tol`: the lower and upper rate-ratio
  equivalence margins;
* `cor_mat`: the dependence between endpoints when there is more than one
  endpoint; and
* `distribution`: either `"pois"` for Poisson counts or `"nbinom"` for
  negative-binomial counts.

The expected count is the rate multiplied by the exposure:

\[
E(Y) = \text{rate} \times \text{exposure}.
\]

For example, a rate of `0.20` per week and an exposure of `10` weeks imply
an expected count of `2` events per participant. Exposure is therefore not
the number of participants. If the supplied rate is already an expected
count per participant over the complete study period, use `exposure = 1`.
The rate and exposure must use compatible time units. Exposure can be a
scalar, an endpoint-specific vector, or arm-specific values when follow-up
differs between arms.

For a Poisson model, the variance equals the mean and no dispersion parameter
is needed. A negative-binomial model is used when the observed counts are more
variable than a Poisson model allows. It requires the additional argument
`dispersion`, which controls the extra variability while leaving the expected
count unchanged:

\[
\operatorname{Var}(Y) = \mu + \text{dispersion}\,\mu^2,
\qquad \mu = \text{rate} \times \text{exposure}.
\]

Thus, with an expected count of `2`, `dispersion = 0.50` gives a variance of
approximately `2 + 0.50 * 2^2 = 4`. Larger dispersion means more variation
between participants and usually less information per participant, which can
increase the required sample size. The dispersion should be chosen from pilot
data or prior knowledge and examined in a sensitivity analysis; it is not a
replacement for the event rate or the exposure.

The remaining planning inputs have the same meaning as for continuous
outcomes: `power`, `alpha`, `nsim`, `seed`, and `dtype` specify the target
power, type-I error level, number of simulations, reproducibility, and study
design, respectively.

## A Published Poisson benchmark

Zhu [@zhu_sample_2017] reports a two-arm Poisson equivalence calculation with
equal event rates of 1 per time unit, average exposure of 0.7, equivalence
limits of 0.9 and $1/0.9$, one-sided $\alpha=0.025$, and 2,705 participants
per arm. The reported total sample size is 5,410, with approximated power
0.80012.

```{r poisson-literature-benchmark}
zhu_benchmark <- simPower(
  n = 2705,
  distribution = "pois",
  rate_list = list(TEST = 1, REF = 1),
  list_comparator = list(TEST_vs_REF = c("TEST", "REF")),
  list_lequi.tol = list(TEST_vs_REF = 0.9),
  list_uequi.tol = list(TEST_vs_REF = 1 / 0.9),
  exposure = 0.7,
  dtype = "parallel",
  alpha = 0.025,
  nsim = 5000,
  seed = 2024
)
zhu_benchmark
data.frame(reference_power = 0.80012,
           simulated_power = zhu_benchmark$power,
           simulated_lower = zhu_benchmark$power_LCI,
           simulated_upper = zhu_benchmark$power_UCI)
```

The simulation need not reproduce the published value exactly. The published
result is a large-sample approximation, whereas `simPower()` simulates
discrete event totals and reports a finite-sample Monte Carlo estimate.

## A Count Study Example

### Step 1:  Define the Study and Theoretical Targets

For this example, we utilize a two-endpoint, two-comparator count study, which will serve as the foundation for the remainder of this vignette. The supplied endpoint correlation is a latent Gaussian-copula
correlation, not the expected Pearson correlation of the observed counts.

```{r poisson-inputs}
count_corr <- matrix(c(1, 0.5, 0.5, 1), nrow = 2,
                     dimnames = list(c("y1", "y2"), c("y1", "y2")))
count_rates <- list(TEST = c(y1 = 0.21, y2 = 0.24),
                    REF = c(y1 = 0.20, y2 = 0.22))
count_comparators <- list(TEST_vs_REF = c("TEST", "REF"))
count_lower <- list(TEST_vs_REF = c(y1 = 0.80, y2 = 0.80))
count_upper <- list(TEST_vs_REF = c(y1 = 1.25, y2 = 1.25))
count_exposure <- 10
```

### Step 2: Run a pilot with sampleSize()
As in the workflow vignette, begin by running a pilot sample size calculation while retaining the simulated data. This object will be used for the diagnostics in the subsequent steps.

```{r poisson-diagnostics}
poisson_diagnostic <- sampleSize(
  distribution = "pois", rate_list = count_rates,
  list_comparator = count_comparators,
  list_lequi.tol = count_lower, list_uequi.tol = count_upper,
  exposure = count_exposure, cor_mat = count_corr,
  dtype = "parallel", nsim = 500, seed = 1234,
  keep_sim_data = TRUE
)
poisson_diagnostic
confint(poisson_diagnostic)
```


### Step 3: Validate the generated data and estimands

The workflow checks result precision, marginal outcomes, exposure-adjusted
rates, endpoint dependence, and Monte Carlo stability before planning.

#### 3a. Arm-specific parameters


```{r poisson-rate-distribution, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(poisson_diagnostic, estimand = "rate")
```

The empirical rate distributions should be centered near the values in
`count_rates`. 

#### 3b. Endpoint correlations
```{r poisson-correlation-distribution, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(poisson_diagnostic, estimand = "correlation",
                  arms = c("TEST", "REF"))
```
The dashed line is the latent correlation supplied through `count_corr`. The
observed Pearson correlation can differ because the latent normal variables
are transformed into discrete counts. The plot checks the direction and
approximate magnitude of the generated dependence; `endpoint_corr` is the
exact parameter check.

#### 3c. Estimand direction and sampling distribution

```{r poisson-outcome-distribution, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(poisson_diagnostic, estimand = "outcome", type = "histogram")
```
The empirical counts should be compatible with the Poisson reference
distribution.

```{r poisson-estimand-distribution, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(poisson_diagnostic, estimand = "RR")
```


### Step 4: Check Monte Carlo stability and precision

```{r poisson-stability, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_stability(poisson_diagnostic)
plot_mc_error(poisson_diagnostic)
```


### Step 5: Check Type I error at the equivalence boundaries
```{r poisson-type1, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
type1_one <- type1Error(x = poisson_diagnostic, null = "both", joint = TRUE)
plot(type1_one)
```


### Step 6: Run the final sample-size calculation
This call uses exactly the rates, exposure, correlation matrix, margins, and
parallel design used by `poisson_diagnostic`.

```{r poisson-sample-size}
poisson_sample_size <- sampleSize(
  power = 0.80, distribution = "pois", rate_list = count_rates,
  list_comparator = count_comparators,
  list_lequi.tol = count_lower, list_uequi.tol = count_upper,
  exposure = count_exposure, cor_mat = count_corr, dtype = "parallel",
  lower = 100, upper = 1000, nsim = 1000, seed = 1234, ncores = 1,
  keep_sim_data = TRUE
)
summary(poisson_sample_size)
confint(poisson_sample_size)
```

For final planning, increase `nsim` and independently verify the selected
sample size with `simPower()`.


```{r poisson-result, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot(poisson_sample_size)
```


### Repeat with a negative-binomial distribution

The negative-binomial calculation keeps the same rates, exposure, correlation,
margins, and parallel design. The dispersion parameter allows the count
variance to exceed its Poisson value. We use `dispersion = 0.50` here as a
deliberately visible sensitivity example. For a mean count near 2, the
Poisson variance is about 2, whereas the negative-binomial variance is
approximately `2 + 0.50 * 2^2 = 4`.

```{r negative-binomial-sample-size}
nb_dispersion <- 0.50
negative_binomial_sample_size <- update(poisson_sample_size,
                                        distribution = "nbinom",
                                        dispersion = nb_dispersion)

summary(negative_binomial_sample_size)
confint(negative_binomial_sample_size)
```

```{r nb-outcome-distribution, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(negative_binomial_sample_size, estimand = "outcome", type = "histogram")
```

The result is conditional on `nb_dispersion`; vary it in sensitivity analyses
when overdispersion is uncertain.

### Use the same settings for a two-by-two crossover

In a crossover calculation, the count-model inputs from the parallel design
are still needed: `rate_list`, `exposure`, `dispersion` (for a
negative-binomial model), `list_comparator`, the equivalence margins, and the
endpoint correlation. These describe the treatment effect and the count
distribution. The crossover design additionally describes how each participant
contributes observations under both treatments:

* `sigmaB` is the between-participant standard deviation on the log-rate
  scale. It represents persistent participant-to-participant heterogeneity in
  event rates.
* `Eper` is a length-two vector of period effects on the log-rate scale. For
  example, `c(0, 0.10)` means that the second period has a 0.10 log-rate
  increase relative to the first period.
* `Eco` is a length-two vector of carry-over effects, ordered as reference
  carry-over and treatment carry-over. Use `c(0, 0)` when there is no
  carry-over effect assumed.
* `dropout` is a length-two vector of dropout proportions for the two
  sequences. Use `c(0, 0)` when no dropout is expected.

For the crossover, `exposure` refers to the observation time or opportunity
for each treatment-period count. A participant contributes one count under
each treatment when complete. For `dtype = "2x2"`, `n` and `n_per_arm` refer
to participants per sequence, not the total number of participants across
both sequences. The crossover-specific parameters should be obtained from
pilot data or subject-matter knowledge; they should not be copied from the
parallel-arm standard deviations.

```{r crossover-negative-binomial-sample-size}

crossover_sample_size <- update(negative_binomial_sample_size,
                                dtype = "2x2",
                                sigmaB = 0.30,
                                Eper = c(0, 0.10),
                                Eco = c(0, 0),
                                dropout = c(0.10, 0.10))

summary(crossover_sample_size)
confint(crossover_sample_size)
```

The crossover calculation uses the same treatment-effect and endpoint
settings but evaluates paired treatment-period data. The values above are
illustrative; in an actual study, `sigmaB`, period effects, carry-over effects,
and sequence-specific dropout should be justified from prior data or a
sensitivity analysis.

## Interpretation and scope

The Poisson, negative-binomial, and crossover results are conditional on the
specified rates, exposure, margins, endpoint dependence, dispersion, and
design. Compare sample sizes only after the corresponding model checks pass.
For count outcomes, `cor_mat` is a latent Gaussian-copula correlation matrix;
it is not a correlation between treatment arms and is not required to equal
the Pearson correlation of the observed counts.
