---
title: "Count Outcomes"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: 'references.bib'
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Count Outcomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE)
library(SimTOST)
```

## Overview

This vignette illustrates the unified `sampleSize()` and `simPower()` interface
for count outcomes. The `distribution` argument selects either a Poisson or a
negative-binomial model, while `dtype` selects a parallel or 2x2 crossover
design. The estimand is the event-rate ratio

\[
RR = \frac{\lambda_T}{\lambda_R},
\]

where \(\lambda_T\) and \(\lambda_R\) are the event rates in the test and
reference groups. Equivalence is assessed with a two-one-sided test on the
log-rate-ratio scale. The examples use equal exposure per participant and
simulation-based power calculations.

The recommended workflow is to separate three questions. First, are the
user-specified rates, exposures, and dispersion values represented correctly in
the simulated data? Second, do the simulated endpoint correlations reproduce
the dependence assumption? Third, what power and sample size follow from that
data-generating model? The first two questions are addressed with the
distribution diagnostics below; they are model checks, not additional
equivalence tests.

## Parallel design: Poisson outcomes

The following example uses event rates of 0.21 and 0.20 per participant for
the test and reference groups, respectively. `simPower()` evaluates the achieved
power at a fixed sample size, while `sampleSize()` searches for the smallest
sample size per arm that reaches the target power.

```{r parallel-poisson-power}
parallel_power <- simPower(
  n = 1000,
  distribution = "pois",
  rate_list = list(TEST = 0.21, REF = 0.20),
  list_comparator = list(TEST_vs_REF = c("TEST", "REF")),
  list_lequi.tol = list(TEST_vs_REF = 0.80),
  list_uequi.tol = list(TEST_vs_REF = 1.25),
  exposure = 1,
  dtype = "parallel",
  nsim = 500,
  seed = 1234,
  keep_sim_data = TRUE
)
parallel_power
confint(parallel_power)
```

```{r parallel-poisson-sample-size}
parallel_sample_size <- sampleSize(
  power = 0.80,
  distribution = "pois",
  rate_list = list(TEST = 0.21, REF = 0.20),
  list_comparator = list(TEST_vs_REF = c("TEST", "REF")),
  list_lequi.tol = list(TEST_vs_REF = 0.80),
  list_uequi.tol = list(TEST_vs_REF = 1.25),
  exposure = 1,
  dtype = "parallel",
  lower = 100,
  upper = 2200,
  nsim = 500,
  seed = 1234
)
parallel_sample_size
summary(parallel_sample_size)
```

The returned sample size is reported per arm and as a total for the two-arm
parallel trial. Simulation error can be reduced by increasing `nsim`.

## Updating a count calculation

Results retain the arguments used to create them, so a new scenario can be
evaluated with `update()` rather than repeating the full call. For example,
the following changes only the fixed sample size and simulation precision for
the power calculation, and only the target power for the sample-size search:

```{r count-update}
parallel_power_larger <- update(parallel_power, n = 1500, nsim = 1000)
parallel_sample_size_90 <- update(parallel_sample_size, power = 0.90,upper = 5000)

parallel_power_larger
summary(parallel_sample_size_90)
```

All other rates, margins, design settings, and random-number settings are
carried forward from the original result. This is useful for sensitivity
analyses, but a changed scientific estimand should still be documented by
showing the relevant endpoint or comparator arguments explicitly.

## A step-by-step validation workflow

The plots in this vignette answer different validation questions. They should
be read in the following order:

1. **Result and precision:** check the estimated power, its Monte Carlo
   interval, and the total sample size.
2. **Marginal outcome model:** check whether simulated counts follow the
   specified Poisson or negative-binomial distribution.
3. **Rate scale:** check whether counts divided by exposure are centered on
   the user-specified event rates.
4. **Endpoint dependence:** check whether within-trial correlations are
   compatible with `cor_mat`.

The first four checks validate the data-generating inputs. Only after those
checks should the power or sample-size conclusion be interpreted.

### Step 1: Check the reported result and Monte Carlo uncertainty

For a fixed sample size, `plot()` shows the achieved power and its confidence
interval. For a sample-size calculation, `summary()` and `confint()` show the
selected sample size and the achieved power at that design. The confidence
interval measures simulation uncertainty; it does not account for uncertainty
in the user-supplied rates, exposure, or dispersion.

```{r count-validation-result, fig.width = 8, fig.height = 5, out.width = "100%", fig.align = "center"}
summary(parallel_sample_size)
confint(parallel_sample_size)
summary(parallel_power)
confint(parallel_power)
plot(parallel_power)
```

Focus on the achieved power and its interval for the fixed-size result. For
the sample-size result, confirm the selected subjects per arm, the resulting
total sample size, and whether achieved power reaches the target. If the
interval is too wide for the planned decision, increase `nsim`; do not treat a
Monte Carlo fluctuation as evidence that the clinical assumptions are wrong.

### Step 2: Check the marginal count distribution

Raw observations are not retained by default. Set `keep_sim_data = TRUE` when
the purpose of the run includes checking the data-generating mechanism. The
following two-endpoint example supplies a latent endpoint correlation of 0.5.
The same parameter can be supplied for a negative-binomial simulation by
changing `distribution`.

```{r poisson-diagnostics}
count_corr <- matrix(c(1, 0.5, 0.5, 1), nrow = 2,
                     dimnames = list(c("y1", "y2"), c("y1", "y2")))
poisson_diagnostic <- simPower(
  n = 500,
  distribution = "pois",
  rate_list = list(TEST = c(y1 = 0.21, y2 = 0.24),
                   REF = c(y1 = 0.20, y2 = 0.22)),
  list_comparator = list(TEST_vs_REF = c("TEST", "REF")),
  list_lequi.tol = list(TEST_vs_REF = c(y1 = 0.80, y2 = 0.80)),
  list_uequi.tol = list(TEST_vs_REF = c(y1 = 1.25, y2 = 1.25)),
  exposure = 1,
  cor_mat = count_corr,
  dtype = "parallel",
  nsim = 500,
  seed = 1234,
  keep_sim_data = TRUE
)
```

`plot_distribution()` with `estimand = "outcome"` shows the distribution of
the individual simulated event counts. For count outcomes, the dashed orange
curve is the specified Poisson or negative-binomial distribution. This checks
the outcome scale, but it is not the most direct check of the rate parameter
because the rate also depends on exposure.

```{r poisson-outcome-distribution, fig.width = 8, fig.height = 5, out.width = "100%", fig.align = "center"}
plot_distribution(poisson_diagnostic, estimand = "outcome", type = "histogram",
                  arms = c("TEST", "REF"))
```

Focus first on the shape and scale of the blue empirical bars. They should be
compatible with the dashed orange Poisson reference. For a negative-binomial
analysis, the same comparison should show the extra variation implied by the
specified dispersion.

### Step 3: Check the exposure-adjusted rate scale

For the parameter-scale check, use `estimand = "rate"`. The function first
combines counts and exposure within each simulated trial and arm, then plots
the resulting exposure-adjusted rate. The dashed line is the rate supplied in
`rate_list`; therefore, the plot directly answers whether the simulated trial
rates are centered on the intended values.

```{r poisson-rate-distribution, fig.width = 8, fig.height = 5, out.width = "100%", fig.align = "center"}
plot_distribution(poisson_diagnostic, estimand = "rate", type = "density",
                  arms = c("TEST", "REF"))
```

Focus on whether the empirical rate distribution is centered near the dashed
line for each arm and endpoint. If it is systematically shifted, check the
rate, exposure, arm names, and endpoint names before interpreting power. This
plot is more informative than a raw count plot when exposure differs between
participants or arms.

### Step 4: Check endpoint dependence

Finally, `plot_distribution(estimand = "correlation")` calculates a correlation between endpoints
within each simulated trial and arm. The blue distribution shows the sampling
variation of those empirical correlations; the dashed orange line is the
latent correlation supplied through `cor_mat`. For discrete outcomes, the
observed Pearson correlation need not equal the latent correlation exactly,
because rates, exposure, overdispersion, and discreteness all affect it.

```{r poisson-correlation-distribution, fig.width = 8, fig.height = 5, out.width = "100%", fig.align = "center"}
plot_distribution(poisson_diagnostic, estimand = "correlation",
                  arms = c("TEST", "REF"))
```

Focus on whether the empirical correlations fluctuate around the dashed
reference value. For count outcomes, the dashed line represents the latent
correlation supplied through `cor_mat`; the observed Pearson correlation can
differ because of discreteness, exposure, event rates, and dispersion.

These diagnostics should be inspected before interpreting power. A mismatch
between the empirical rate or correlation distributions and their reference
values usually indicates that the input structure, exposure specification, or
simulation assumptions need review.

After these four checks, repeat the workflow for the negative-binomial and
2x2 crossover scenarios if they are plausible alternatives for the study. A
model that gives the desired power but fails one of these input checks should
be corrected before changing the sample size or equivalence margins.

## Published two-arm Poisson benchmark

As a simple external check, Zhu [@zhu_sample_2017] reports a two-arm Poisson
equivalence calculation with equal event rates of 1 per time unit, average
exposure of 0.7, equivalence limits of 0.9 and $1/0.9$, one-sided
\\(\\alpha=0.025\\), and 2,705 participants per arm. The reported total sample
size is therefore 5,410, with approximated power 0.80012. The following uses
the unified `simPower()` interface for the same treatment-versus-control
scenario.

```{r poisson-literature-benchmark}
zhu_benchmark <- simPower(
  n = 2705,
  distribution = "pois",
  rate_list = list(TEST = 1, REF = 1),
  list_comparator = list(TEST_vs_REF = c("TEST", "REF")),
  list_lequi.tol = list(TEST_vs_REF = 0.9),
  list_uequi.tol = list(TEST_vs_REF = 1 / 0.9),
  exposure = 0.7,
  dtype = "parallel",
  alpha = 0.025,
  nsim = 5000,
  seed = 2024
)

zhu_benchmark
data.frame(
  reference_power = 0.80012,
  simulated_power = zhu_benchmark$power,
  simulated_lower = zhu_benchmark$power_LCI,
  simulated_upper = zhu_benchmark$power_UCI
)
```

The simulation should be close to the published approximation, but it is not
expected to reproduce it exactly. Zhu's value is based on a large-sample
calculation, whereas `simPower()` simulates discrete Poisson event totals and
uses the package's finite-sample equivalence procedure. In addition, the
Monte Carlo estimate varies slightly with `nsim` and the random seed. The
benchmark is intended as a plausibility check for the simple two-arm case;
the three-arm, multi-endpoint examples below address a different, joint
multiple-comparison problem.

Exposure and negative-binomial dispersion can also vary by arm. Each list must
use the same arm names as `rate_list`; values may be scalar or endpoint
vectors.

```{r arm-specific-count-parameters, eval = FALSE}
simPower(
  n = 1000,
  distribution = "nbinom",
  rate_list = list(TEST = c(endpoint_1 = 0.21),
                   REF = c(endpoint_1 = 0.20)),
  list_comparator = list(TEST_vs_REF = c("TEST", "REF")),
  list_lequi.tol = list(TEST_vs_REF = 0.80),
  list_uequi.tol = list(TEST_vs_REF = 1.25),
  exposure = list(TEST = 1, REF = 2),
  dispersion = list(TEST = 0.005, REF = 0.010),
  dtype = "parallel", nsim = 1000, seed = 1234
)
```

## 2x2 crossover design: negative-binomial outcomes

The same unified functions can plan a balanced 2x2 crossover design. For
`dtype = "2x2"`, `n` in `simPower()` and the returned `n_per_arm` in
`sampleSize()` refer to the number of participants per sequence. The
negative-binomial example uses arm-specific exposure and dispersion values;
larger dispersion values represent more overdispersion.

```{r crossover-negative-binomial-power}
crossover_power <- simPower(
  n = 1000,
  distribution = "nbinom",
  rate_list = list(TEST = 0.21, REF = 0.20),
  list_comparator = list(TEST_vs_REF = c("TEST", "REF")),
  list_lequi.tol = list(TEST_vs_REF = 0.80),
  list_uequi.tol = list(TEST_vs_REF = 1.25),
  exposure = list(TEST = 1, REF = 1.2),
  dispersion = list(TEST = 0.005, REF = 0.008),
  dtype = "2x2",
  nsim = 200,
  seed = 1234
)
crossover_power
confint(crossover_power)
```

```{r crossover-negative-binomial-sample-size}
crossover_sample_size <- sampleSize(
  power = 0.80,
  distribution = "nbinom",
  rate_list = list(TEST = 0.21, REF = 0.20),
  list_comparator = list(TEST_vs_REF = c("TEST", "REF")),
  list_lequi.tol = list(TEST_vs_REF = 0.80),
  list_uequi.tol = list(TEST_vs_REF = 1.25),
  exposure = list(TEST = 1, REF = 1.2),
  dispersion = list(TEST = 0.005, REF = 0.008),
  dtype = "2x2",
  lower = 300,
  upper = 1200,
  nsim = 100,
  seed = 1234
)
crossover_sample_size
summary(crossover_sample_size)
```

In the current count crossover implementation, the two sequences are
balanced and each participant contributes one exposure under each treatment.
The kernel can additionally accept `sigmaB`, `Eper`, `Eco`, and `dropout` to
represent between-subject variability, period effects, carry-over effects,
and sequence-specific dropout. The paired log-rate analysis and its
assumptions are described in the companion
`methodological_assumptions.Rmd` vignette.

## Interpretation and scope

The count extension is intended for planning studies in which the estimand is
an event-rate ratio based on aggregate event counts. Users should select the
Poisson or negative-binomial model using prior information about event-rate
variability. The simulated power is conditional on the specified rates,
exposure, equivalence margins, dispersion, and design assumptions.

For multi-endpoint count simulations, `cor_mat` is a latent Gaussian-copula
correlation matrix. It describes dependence between endpoints within the
simulation, not the Pearson correlation of the observed discrete counts and
not a correlation between the test and reference arms. Use `diag(m)` for
latent endpoint independence. Because the observed correlation also depends
on the rates, exposure, dispersion, and discreteness of the counts, `cor_mat`
should be justified from prior data or varied in sensitivity analyses.

The diagnostics above make this distinction visible: `plot_distribution()`
checks the marginal outcome or rate distributions, whereas
`plot_distribution(estimand = "correlation")` checks the dependence generated between endpoints. Neither
plot estimates an additional treatment effect; the estimand used for the
equivalence decision remains the event-rate ratio
\[
RR = \lambda_T / \lambda_R.
\]
