---
title: "Methodological Assumptions in SimTOST"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methodological Assumptions in SimTOST}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE,
                      comment = "#>", warning = FALSE, message = FALSE)
library(SimTOST)
```

## Purpose and scope

SimTOST estimates sample size by repeatedly simulating complete studies and
applying the planned equivalence procedure to each simulated study. The same
framework is available through `sampleSize()` for planning and `simPower()`
for evaluating power at a fixed sample size. The `distribution` argument
selects the outcome model:

```{r distributions}
c("norm", "lnorm", "pois", "nbinom")
```

This vignette describes the assumptions that should be considered when using
these models. It is methodological guidance, not a substitute for a
prespecified statistical analysis plan.

## Assumptions shared by the simulation framework

The following assumptions apply to all outcome distributions unless explicitly
changed by the design inputs:

* treatment arms, endpoint definitions, exposure periods, and equivalence
  margins are specified before the simulation;
* the supplied means, rates, variances, correlations, and dispersion values
  represent the intended study population;
* observations are generated according to the selected marginal outcome model;
* dropout is represented through the supplied arm- or sequence-specific
  dropout proportions and is non-informative with respect to the outcome;
* the selected design (`dtype = "parallel"` or `dtype = "2x2"`) matches the
  planned trial;
* the number of simulations is sufficiently large for the desired Monte Carlo
  precision; and
* the equivalence margins, endpoint rule (`k`), and multiplicity adjustment
  are the actual decision rules that will be used in the trial.

For a fixed number of simulations, the reported power is a Monte Carlo
estimate. Its confidence interval describes simulation uncertainty; it does
not describe uncertainty in the assumed rates, means, variances, or
correlations.

## Continuous outcomes

### Normal outcomes

For `distribution = "norm"`, the simulated endpoint vector is multivariate
normal within each treatment arm. The supplied `mu_list` contains arm- and
endpoint-specific means. Variability is supplied through `varcov_list`, or is
constructed from `sigma_list` and `cor_mat` (or the common correlation `rho`).

The Normal model assumes:

* continuous outcomes on the analysis scale;
* approximately symmetric marginal distributions;
* a common covariance structure within each arm for the simulated study;
* a valid, positive-definite covariance matrix; and
* no unmodelled skewness or heavy tails that would materially alter the TOST
  statistic.

The DOM test uses additive equivalence margins. The ROM test is appropriate
when the scientific question concerns a ratio of means and the outcome scale
supports that interpretation.

### Log Normal outcomes

For `distribution = "lnorm"`, the simulated outcomes are positive and
right-skewed. The ratio-of-means procedure is applied on the scale specified by
the package implementation and the supplied means and standard deviations
must be interpreted consistently with that implementation.

The Log Normal model assumes:

* strictly positive outcomes;
* approximately normal log-outcomes;
* endpoint dependence represented through the supplied covariance or
  correlation inputs; and
* a ratio-based equivalence question rather than an additive one.

The Log Normal model should not be used for outcomes with structural zeros or
negative values without an explicit transformation and a corresponding
reconsideration of the estimand.

`distribution = "lnorm"` with `ctype = "DOM"` is rejected because the
current implementation converts arithmetic means and covariances to the log
scale and applies a DOM test there. That has a ratio interpretation on the
original scale, not an additive difference interpretation. Use
`distribution = "lnorm", ctype = "ROM"` for an arithmetic mean ratio,
or use the normal distribution for an additive mean difference.

## Count outcomes

For count outcomes, `rate_list` supplies the event rate per unit exposure.
For a parallel arm $a$ and endpoint $j$, the expected aggregate count is

\[
  \mathrm{E}(Y_{aj}) = n_a e_{aj}\lambda_{aj},
\]

where $n_a$ is the number of participants, $e_{aj}$ is exposure, and
$\lambda_{aj}$ is the event rate. Exposure can be scalar, endpoint-specific,
or arm-specific. Count equivalence is assessed through a rate ratio and
log-rate-ratio TOST.

### Poisson outcomes

For `distribution = "pois"`, each marginal count follows a Poisson model:

\[
  Y_{aj} \sim \operatorname{Poisson}(n_a e_{aj}\lambda_{aj}).
\]

The Poisson assumption implies that the variance equals the mean. This is
appropriate only when additional heterogeneity, clustering, and exposure
variation are negligible or have already been incorporated into the model.
Overdispersion caused by unobserved subject heterogeneity or recurrent-event
dependence can make a Poisson analysis anticonservative.

### Negative Binomial outcomes

For `distribution = "nbinom"`, let $\phi_{aj} > 0$ denote the per-subject
dispersion parameter. A subject-level count with mean

\[
\mu^{(\mathrm{subj})}_{aj} = e_{aj}\lambda_{aj}
\]

has variance

\[
\operatorname{Var}(Y^{(\mathrm{subj})}_{aj})
 = \mu^{(\mathrm{subj})}_{aj}
 + \phi_{aj}\left(\mu^{(\mathrm{subj})}_{aj}\right)^2,
\]

which is the negative-binomial parameterization with
$\operatorname{size}=1/\phi_{aj}$. In a parallel design, the aggregate count
over $n_a$ independent subjects has mean

\[
\mu_{aj} = n_a e_{aj}\lambda_{aj}
\]

and the implementation uses
$\operatorname{size}=n_a/\phi_{aj}$. Equivalently, its variance is

\[
\operatorname{Var}(Y_{aj})
 = n_a\left\{\mu^{(\mathrm{subj})}_{aj}
 + \phi_{aj}\left(\mu^{(\mathrm{subj})}_{aj}\right)^2\right\}.
\]

Thus, larger values of `dispersion` produce more variation beyond the
Poisson variance, while values closer to zero give behavior closer to the
Poisson model. This parallel-design scaling of the size parameter is
important: using $1/\phi_{aj}$ for an aggregate count would incorrectly make
the overdispersion increase with sample size. The dispersion parameter is
positive and should preferably be based on historical data, pilot data, or a
clinically justified sensitivity range.

The Negative Binomial model assumes that the selected mean-dispersion
relationship adequately represents overdispersion in the planned study.

## Dependence between endpoints

### Continuous outcomes

For continuous outcomes, endpoint dependence is represented directly through
the covariance matrices in `varcov_list`, or through `sigma_list` together
with `cor_mat` or `rho`. The covariance matrices must be compatible with the
endpoint means and standard deviations and must be positive definite.

### Count outcomes and the Gaussian copula

For joint count simulations, `cor_mat` is interpreted as the correlation
matrix of latent Gaussian variables. For each arm, SimTOST:

1. draws a multivariate standard-normal latent vector with correlation
   `cor_mat`;
2. transforms each latent value to a uniform value using the standard normal
   CDF; and
3. transforms each uniform value through the Poisson or negative-binomial
   inverse CDF for the specified arm and endpoint.

This is a Gaussian-copula construction. It preserves the selected marginal
count distributions while inducing dependence between endpoints. The entries
of `cor_mat` are not Pearson correlations of the observed counts. Observed
count correlations also depend on rates, exposure, and dispersion, and
discreteness means that the raw-count correlation need not equal the supplied
latent correlation.

#### What `cor_mat` means in practice

For endpoints \(j=1,\ldots,m\), the user-supplied matrix
\(\boldsymbol{R}=\texttt{cor\_mat}\) is the correlation matrix of a latent
standard-normal vector

\[
\boldsymbol{Z}_a \sim N_m(\boldsymbol{0},\boldsymbol{R})
\]

for arm \(a\). Each component is transformed as

\[
U_{aj}=\Phi(Z_{aj}), \qquad
Y_{aj}=F^{-1}_{aj}(U_{aj}),
\]

where \(F_{aj}\) is the specified Poisson or negative-binomial marginal
distribution. Thus, `cor_mat[1, 2] = 0.8` means that endpoints 1 and 2 have
latent Gaussian correlation 0.8 before they are transformed into counts. It
does not mean that their observed event counts will have Pearson correlation
0.8.

The matrix has the following interpretation:

* diagonal elements must be 1;
* off-diagonal elements describe dependence between endpoints within the same
  arm or participant-level simulation unit;
* the matrix does not describe dependence between test and reference arms;
* it does not describe dependence between treatment comparisons; and
* it does not replace the subject-level pairing used by the 2x2 crossover
  analysis.

An identity matrix, `diag(m)`, gives independent latent endpoint simulations.
Positive correlations generally increase the probability that endpoint tests
pass or fail together, which can materially change the probability of meeting
the `k`-endpoint rule. The matrix should therefore be based on historical or
pilot information, or varied in sensitivity analyses. It must be symmetric,
positive definite, and have unit diagonal.

The latent vectors are generated independently between arms. In a three-arm
study, however, the same simulated test-arm outcomes are used in the
test-versus-reference comparisons. Thus, joint power reflects both endpoint
dependence and the shared test arm. An identity matrix, `diag(m)`, represents
latent endpoint independence. In a 2x2 crossover, the same copula is applied
to endpoint-specific subject effects and to the period-specific count
innovations within a participant. This preserves the intended endpoint
dependence while retaining the within-participant pairing.

## Design-specific assumptions

### Parallel design

For a parallel design, observations are generated independently between
participants and treatment arms, apart from endpoint dependence specified by
the model. Allocation and dropout determine the number of analyzable
participants in each arm.


### 2x2 crossover design

#### Normal and Log-Normal outcomes

For continuous outcomes, `dtype = "2x2"` represents a balanced two-sequence,
two-period crossover. The two treatment sequences are reference--test (RT)
and test--reference (TR), and `n` is interpreted as the number of subjects per
sequence before dropout. On the analysis scale, the data-generating model can
be written for subject i, period p, and endpoint j as

\[
  Y_{ipj} = \mu_{a(i,p),j} + E_p + C_{a(i,p),s(i)} + b_i +
  \varepsilon_{ipj},
\]

where a(i,p) is the treatment received, E_p is the period effect,
C(a,s) is the carry-over effect determined by the preceding treatment,
b_i is a subject-level effect, and the vector of
within-subject errors has covariance matrix `SigmaW`. The same subject effect
is used for both periods, which induces within-subject dependence and is
important for the crossover comparison. `sigmaB` is the between-subject
standard-deviation parameter on the scale used by the simulation kernel;
`SigmaW` describes residual within-subject endpoint variability and
correlation.

For `distribution = "norm"` with `ctype = "DOM"`, the simulated outcomes
are generated directly on the supplied continuous analysis scale. The
equivalence test compares the treatment--reference difference with additive
equivalence limits. The crossover kernel estimates treatment and reference
means by averaging the corresponding sequence-period means and uses the
within-subject variation in the TOST standard error.

For `distribution = "lnorm"`, the supplied arithmetic means and standard
deviations are converted to the corresponding log-scale means and covariance
matrix before simulation. The equivalence limits are also transformed by the
log function, and the DOM kernel is then applied on the log scale. Consequently,
the resulting decision has a ratio interpretation on the original scale. A
Log-Normal outcome must be strictly positive; structural zeros or negative
values require a different model or a prespecified transformation.

The continuous 2x2 implementation assumes that:

* both sequences contain subjects observed in both periods after the supplied
  sequence-level dropout adjustment;
* treatment, period, and carry-over effects are additive on the analysis
  scale, with no treatment-by-period interaction unless represented through
  the supplied inputs;
* the within-subject covariance matrix is valid and positive definite;
* the subject effect is adequately represented by a normal random effect and
  is independent of the residual errors; and
* dropout is non-informative and reduces the available subjects by sequence;
  period-specific missing observations are not modelled by this continuous
  kernel.

`Eper = c(E_1, E_2)` specifies the two period effects. `Eco` is ordered as
`c(reference_carryover, treatment_carryover)`: in the RT sequence, treatment
in period 2 receives the reference carry-over effect, whereas in the TR
sequence, reference in period 2 receives the treatment carry-over effect.
Under balanced sequences and no differential carry-over, the treatment effect
is identified after averaging the two sequence-specific estimates and the
period effect cancels. Non-zero carry-over values change the simulated means
and therefore change the estimand being evaluated; they should be prespecified
and subjected to sensitivity analysis.

Endpoint dependence for these continuous crossover simulations is supplied
directly through `SigmaW` (or through `sigma_list` and `cor_mat` when the
covariance matrix is constructed). This differs from the count implementation,
where `cor_mat` is used as a latent Gaussian-copula correlation.

#### Poisson and Negative Binomial outcomes

For a 2x2 crossover design, the count extension uses a paired log-link rate
analysis. Each complete participant contributes one count under the test
treatment and one count under the reference treatment. The count-generating
model is, conditionally on the subject effect,

\[
Y_{isp} \sim \begin{cases}
\operatorname{Poisson}(e_{isp}\lambda_{isp}), &\text{Poisson},\\
\operatorname{NegBin}(e_{isp}\lambda_{isp},\phi_{isp}),
  &\text{Negative Binomial},
\end{cases}
\]

with

\[
\log(\lambda_{isp}) = \log(\lambda_{i}) + E_{p} + C_{isp} + b_{is},
\qquad b_{is}\sim N(0,\sigma_B^2).
\]

Here, `exposure` is the offset \(e_{isp}\), `Eper = c(E_1,E_2)` contains
period effects, `Eco = c(C_R,C_T)` contains reference- and treatment-
carry-over effects, and `sigmaB` is the standard deviation of the subject
random intercept on the log-rate scale. The negative-binomial dispersion is
used in the subject-period count generation through the package's
size-parameter convention \(\operatorname{size}=1/\phi\).

The analysis forms a log-rate contrast within each complete participant,
averages the contrasts within each sequence, and then averages the two
sequence estimates. This removes the period effect under a balanced 2x2
design. The analysis applies the corresponding carry-over correction using
`Eco`; therefore, nonzero carry-over values change both the generated counts
and the estimand adjustment. Exposure enters the contrast as an offset rather
than being treated as an additional treatment effect. The standard error is
estimated from the empirical variance of the subject-level contrasts, so the
Poisson or negative-binomial sampling variation and the effect of the
specified dispersion are retained.

The within-subject contrast removes the subject random intercept from the
treatment effect. Consequently, `sigmaB` affects the simulated paired counts
but should not be interpreted as an additional treatment-effect variance in
the paired contrast. This is a conditional, paired rate-ratio analysis; it is
not a marginal population-average model and it does not estimate an
independent carry-over coefficient. If substantial carry-over is expected,
the design and estimand should be reconsidered because correction relies on
the supplied `Eco` values.

Sequence-specific dropout is applied before the paired analysis. Participants
missing either period are excluded from the within-subject contrast. The
method therefore assumes non-informative dropout and requires enough complete
participants in both sequences. A crossover design should not be used merely
as a computational substitute for a parallel design.

For multiple endpoints in either design, the endpoint-specific analyses are
combined using `k`, while the Gaussian copula determines the joint simulated
success event. Thus, `cor_mat` is used by both the joint parallel engine and
the multi-endpoint 2x2 engine. The supplied matrix remains a latent Gaussian
correlation, not necessarily the Pearson correlation of the observed counts.

## Practical sensitivity analyses

Because power depends on assumptions that are rarely known exactly, planning
should examine sensitivity to:

* event rates or continuous-outcome means;
* exposure and dropout;
* standard deviations or covariance matrices;
* negative-binomial dispersion;
* endpoint correlations or the latent count copula correlation;
* the number of required endpoints `k`; and
* the multiplicity adjustment.

The selected scenario should be justified in the statistical analysis plan,
and the final reported sample size should account for Monte Carlo uncertainty
and any operational inflation required by the study.
