---
title: "Bioequivalence Tests for Parallel Trial Designs: 2 Arms, 3 Endpoints"
author: "Thomas Debray"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    fig_caption: yes
    fig_width: 9
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Bioequivalence Tests for Parallel Trial Designs: 2 Arms, 3 Endpoints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: 'references.bib'
link-citations: yes
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
options(rmarkdown.html_vignette.check_title = FALSE) #title of doc does not match vignette title
doc.cache <- T #for cran; change to F
```

# Introduction

In many studies, it is necessary to evaluate equivalence across multiple primary variables. For instance, the European Medicines Agency (EMA) recommends demonstrating equivalence for both **Area Under the Curve** (AUC) and **maximum concentration** (Cmax) when assessing pharmacokinetic properties.

When multiple primary endpoints are involved, a decision must be made on the desired criteria for equivalence:

* Equivalence for All Primary Endpoints
  * This is the most common setting and is often referred to as having *multiple co-primary endpoints*.
  * Equivalence must be demonstrated for **all** endpoints to conclude overall equivalence.
* Equivalence for At Least One Primary Endpoint
  * Known as having *multiple primary endpoints*.
  * Equivalence is required for **at least one** endpoint to meet the study's objectives.

This vignette presents advanced techniques for calculating sample size in parallel trial designs involving three treatment arms and two endpoints. Specifically, it focuses on bioequivalence testing between a new treatment (SB2) and a reference drug (Remicade) administered in two distinct locations (EU_Remicade and USA_Remicade).

As an illustrative example, we consider published data from the phase-1 trial [NCT01922336](https://clinicaltrials.gov/study/NCT01922336#study-overview). This trial assessed the pharmacokinetics of SB2 compared to its EU-sourced reference product. The following outcomes were reported following a single dose of SB2 or its EU reference product [@shin_randomized_2015]:

```{r, echo=FALSE}
data <- data.frame("PK measure" = c("AUCinf ($\\mu$g*h/mL)","AUClast ($\\mu$g*h/mL)","Cmax ($\\mu$g/mL)"),
                   "SB2" = c("38,703 $\\pm$ 11,114", "36,862 $\\pm$ 9133", "127.0 $\\pm$ 16.9"), 
                   "EU-INF" = c("39,360  $\\pm$ 12,332", "37,022 $\\pm$ 9398", "126.2 $\\pm$ 17.9"))

kableExtra::kable_styling(kableExtra::kable(data, 
                                            col.names = c("PK measure", "SB2", "Remicade (EU)"),
                                            caption = "Primary PK measures between test and reference product. Data represent arithmetic mean +- standard deviation."),
                          bootstrap_options = "striped")
```

# Methodology and Assumptions

The bioequivalence analysis focuses on two key pharmacokinetic endpoints:

* AUCinf: Area Under the Curve (infinity)
* Cmax: Maximum concentration

For both endpoints, the analysis assumes that:

* Summary data (e.g., mean and standard deviation) are available on the original scale.
* These data are provided for each treatment arm.

To evaluate bioequivalence, we apply the $80\%/125\%$ rule, which defines equivalence bounds relative to the reference mean. The evaluation is conducted using a one-sided significance level of 5\%, with a target statistical power of 90\%.

# Testing the Difference of Means (DOM) for multiple Co-primary Endpoints
In biosimilar development, it is important to demonstrate equivalence across all relevant doses, routes of administration, patient populations, and endpoints. To establish equivalence between two treatments, the difference in means for each endpoint, $\mu_{T}^{(j)} - \mu_{R}^{(j)}$, must lie within a predefined equivalence margin around zero for all primary endpoints.

## Hypotheses
The null and alternative hypotheses for the equivalence test are as follows:

Null Hypothesis ($H_0$): At least one endpoint does not meet the equivalence criteria:
 
$$H_0: \mu_T^{(j)} - \mu_R^{(j)} \le E_L ~~ \text{or}~~ \mu_T^{(j)} - \mu_R^{(j)} \ge E_U \quad \text{for at least one}\;j$$

Alternative Hypothesis ($H_1$): All endpoints meet the equivalence criteria:

$$H_1: E_L<\mu_{T}^{(j)}-\mu_{R}^{(j)} < E_U \quad\text{for all}\;j$$

The null hypothesis ($H_0$) is rejected if, and only if, all null hypotheses associated with the $K$ primary endpoints are rejected at a significance level of $\alpha$. This ensures that equivalence is established across all endpoints simultaneously.

## Statistical Considerations

* **Type I Error Control**: Since rejection of $H_0$ requires all individual null hypotheses to be rejected, there is no need for multiplicity adjustments. The Type I error rate is controlled by the design.
* **Impact on Power**: Requiring equivalence across multiple endpoints decreases the overall power of the test. The Type II error increases as the number of primary endpoints ($K$) grows, which can make equivalence testing more challenging [@mielke_sample_2018].

## Independent Testing of Pharmacokinetic (PK) Measures
If each pharmacokinetic (PK) measure is tested independently, the following sample sizes would be required for each endpoint to achieve a 5\% significance level:

```{r}
library(SimTOST)

# Sample size calculation for AUCinf
(sim_AUCinf <- sampleSize(
  power = 0.9,                                # Target power
  alpha = 0.05,                               # Significance level
  arm_names = c("SB2", "EU_Remicade"),        # Names of trial arms
  list_comparator = list("EMA" = c("SB2", "EU_Remicade")),  # Comparator configuration
  mu_list = list("SB2" = 38703, "EU_Remicade" = 39360),     # Mean values
  sigma_list = list("SB2" = 11114, "EU_Remicade" = 12332),  # Standard deviation values
  list_lequi.tol = list("EMA" = 0.80),        # Lower equivalence margin
  list_uequi.tol = list("EMA" = 1.25),        # Upper equivalence margin
  ncores = 1,                                 # Number of computation cores
  nsim = 1000                                 # Number of stochastic simulations
))

# Sample size calculation for AUClast
(sim_AUClast <- sampleSize(
  power = 0.9,                                # Target power
  alpha = 0.05,                               # Significance level
  arm_names = c("SB2", "EU_Remicade"),        # Names of trial arms
  list_comparator = list("EMA" = c("SB2", "EU_Remicade")),  # Comparator configuration
  mu_list = list("SB2" = 36862, "EU_Remicade" = 37022),     # Mean values
  sigma_list = list("SB2" = 9133, "EU_Remicade" = 9398),    # Standard deviation values
  list_lequi.tol = list("EMA" = 0.80),        # Lower equivalence margin
  list_uequi.tol = list("EMA" = 1.25),        # Upper equivalence margin
  ncores = 1,                                 # Number of computation cores
  nsim = 1000                                 # Number of stochastic simulations
))
```


If we were to test each PK measure independently, we would find a total sample size of `r sim_AUCinf$response$n_total` for AUCinf, `r sim_AUClast$response$n_total` for AUClast, and $N=20$ for Cmax. This means that we would have to enroll `r sim_AUCinf$response$n_total` + `r sim_AUClast$response$n_total` +$20 = 152$ patients in order to reject $H_0$ at a significance level of 5\%.



# References
