---
title: "A Step-by-Step Simulation Validation Workflow"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Step-by-Step Simulation Validation Workflow}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: 'references.bib'
link-citations: yes
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", collapse = TRUE)
options(rmarkdown.html_vignette.check_title = FALSE)
```

This vignette shows how to validate a simulation before interpreting its
sample-size or power result. The example has three arms (`T`, `R1`, and
`R2`), three correlated endpoints, and two comparisons: `R1` versus `T`
and `R2` versus `T`.

The important principle is that the final power calculation is the last
step, not the first. A power result can be numerically precise while
still being wrong if the data-generating model, estimand direction,
correlation, or decision rule was specified incorrectly.

The workflow therefore follows this order:

1.  define the intended study and its theoretical targets;
2.  run a pilot sample-size calculation with moderate `nsim`;
3.  validate the generated outcomes, parameters, estimands, and
    correlations;
4.  check whether the number of simulation trials is sufficient;
5.  check Type I error at the equivalence boundaries;
6.  run the final sample-size calculation;
7.  interpret the final result and inspect endpoint-level decisions.

If a check fails, stop and correct the inputs or implementation before
proceeding. Increasing `nsim` only reduces Monte Carlo noise; it does
not correct a wrong mean, distribution, estimand direction, correlation,
or equivalence margin.

```{r setup-package, message=FALSE}
library(SimTOST)
```

## Step 1: Define the study and theoretical targets

The estimand is the ratio of means (ROM). For every endpoint and
comparison, the equivalence hypotheses are

$$H_0: \mathrm{ROM} \le L \text{ or } \mathrm{ROM} \ge U,
\qquad
H_1: L < \mathrm{ROM} < U.$$ The first arm in each comparator is the
test arm and the second is the reference arm. Thus `T_vs_R1` means
`T / R1`, and `T_vs_R2` means `T / R2`. The equivalence limits are 0.80
and 1.25, and `rho = 0.50` is the intended within-arm correlation
between endpoints.

```{r workflow-inputs}
endpoints <- paste0("y", 1:3)
mu_t <- setNames(rep(1.00, 3), endpoints)
mu_r1 <- setNames(rep(1.05, 3), endpoints)
mu_r2 <- setNames(rep(1.04, 3), endpoints)
sigma_t <- setNames(rep(0.20, 3), endpoints)
sigma_r1 <- setNames(rep(0.20, 3), endpoints)
sigma_r2 <- setNames(rep(0.21, 3), endpoints)

comparators <- list(T_vs_R1 = c("T", "R1"), T_vs_R2 = c("T", "R2"))
lower <- list(T_vs_R1 = setNames(rep(0.80, 3), endpoints),
              T_vs_R2 = setNames(rep(0.80, 3), endpoints))
upper <- list(T_vs_R1 = setNames(rep(1.25, 3), endpoints),
              T_vs_R2 = setNames(rep(1.25, 3), endpoints))
```

Before simulating, check that every comparator uses the intended
direction, that the endpoint names match in both arms, and that the
margins match the scientific question. If any of these are wrong,
correct them now. Later diagnostic plots cannot identify a wrongly
specified scientific target.

## Step 2: Run a pilot with `sampleSize()`

Here the pilot is the sample-size calculation itself. It searches for a
candidate design using a moderate `nsim` and retains the simulated
trials at the selected candidate. `keep_sim_data = TRUE` is required for
the distribution, parameter, estimand, and correlation diagnostics.

```{r workflow-pilot, eval=TRUE}
ss_pilot <- sampleSize(
  power = 0.80, alpha = 0.05,
  mu_list = list(T = mu_t, R1 = mu_r1, R2 = mu_r2),
  sigma_list = list(T = sigma_t, R1 = sigma_r1, R2 = sigma_r2),
  rho = 0.50, list_comparator = comparators,
  list_y_comparator = list(T_vs_R1 = endpoints, T_vs_R2 = endpoints),
  list_lequi.tol = lower, list_uequi.tol = upper,
  dtype = "parallel", adjust = "no", k = 3,
  distribution = "lnorm", ctype = "ROM", ncores = 1,
  nsim = 500, seed = 2026, keep_sim_data = TRUE
)
```

The returned `simss` object is the `k = 3` pilot used for the diagnostic
checks below. To evaluate the same pilot design under the alternative
rule that requires at least two of the three endpoints per comparator,
update the pilot. `update()` retains the design, endpoint, margin, and
simulation settings and replaces only the supplied arguments:

```{r workflow-pilot-k2, eval=TRUE}
ss_pilot_k2 <- update(ss_pilot, adjust = "bon", k = 2, seed = 2026)
```

`ss_pilot_k2` is a separate `simss` object: its selected sample size and
achieved power are based on the 2-of-3 decision rule and should not be
confused with the `k = 3` pilot. This illustrative call deliberately
uses the conservative Bonferroni adjustment due to Type I error
inflation. For completeness, the same pilot design can also be evaluated
under the rule that at least one of the three endpoints must pass for
each comparator. This is another update of the pilot, with `k = 1`:

```{r workflow-pilot-k1, eval=TRUE}
ss_pilot_k1 <- update(ss_pilot, adjust = "bon", k = 1, seed = 2027)
```

## Step 3: Validate the generated data and estimands

These checks answer: “Did the simulator generate the model that I
specified?” They are performed before the final power calculation.

The simplified diagnostic interface is
`plot_distribution(x, estimand = ..., arms = ..., endpoints = ...)`.
The `estimand` can be `"mu"`, `"sigma"`, `"t_value"`, `"DOM"`,
`"ROM"`, `"RR"`, or `"correlation"`. If `arms` or `endpoints` is omitted,
all available values are displayed. The optional `type` argument controls
whether the result is shown as a density, histogram, or ECDF (and Q--Q plot
where applicable).

### 3a. Arm-specific parameters

The following plots compare trial-level means and standard deviations
with the user-specified values. The finite-sample estimates will vary
around the dashed line; the key question is whether they are centered
correctly.

```{r workflow-parameter-mean, eval=TRUE, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(ss_pilot, estimand = "mu")
```

```{r workflow-parameter-sigma, eval=TRUE, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(ss_pilot, estimand = "sigma")
```

If the distributions are centered away from their references, check the
input parameterization and the simulation code. If they are very wide,
that may be expected for the pilot sample size; it is a precision issue,
not necessarily a bias issue.

### 3b. Endpoint correlations

This plot checks whether the empirical within-arm endpoint correlations
are centered near the specified value of 0.50. It checks dependence
between endpoints in the same arm, not dependence between treatment
arms.

```{r workflow-correlation, eval=TRUE, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(ss_pilot, estimand = "correlation")
```

If the correlations are wrong, check `rho` or `cor_mat`, endpoint order,
and the covariance construction. Do not interpret joint power until this
check is satisfactory because endpoint correlation affects the
probability that multiple endpoint tests pass together.

### 3c. Estimand direction and sampling distribution

This plot checks the quantity used by the test. Because the comparators
are `c(test, reference)`, the theoretical ROM values are approximately
0.95 for `T_vs_R1` and 0.96 for `T_vs_R2`.

```{r workflow-estimand, eval=TRUE, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(ss_pilot, estimand = "ROM")
```

If the distribution is centered near the reciprocal, such as
approximately 0.95 instead of 1.05, the comparator direction has been
reversed. Correct `list_comparator` before continuing. A shift that is
not a reciprocal usually indicates an issue with the supplied means or
the ROM calculation.


### 3d. TOST t-statistics

You can also inspect the distributions of the two one-sided test
statistics used by the TOST procedure:

```{r workflow-t-values, eval=TRUE, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_distribution(ss_pilot, estimand = "t_value",
                  arms = c("T", "R1", "R2"))
```

Each panel is one endpoint and comparator. The blue and orange curves
are the lower-bound and upper-bound t-statistics. The dashed lines are
their one-sided critical values. For an endpoint to pass equivalence,
the lower statistic should be to the right of its critical line and the
upper statistic should be to the left of its critical line. If one curve
frequently lies on the wrong side, that boundary is responsible for many
endpoint failures.

These t-statistics are reconstructed from the retained outcomes using
the same continuous parallel-test formulas as the TOST calculation. They
are a diagnostic of the statistic's distribution, not a replacement for
the trial-level decisions in `plot_decision_heatmap()` or the numerical
power in `summary(ss)`.

The `n_trials` argument used in the plots only limits the observations
displayed. It is not the number of Monte Carlo trials used by the power
calculation; that number is `nsim`.

## Step 4: Check Monte Carlo stability and precision

This step addresses: *"Is the number of simulations (nsim) sufficient
for a stable estimate?"* It assesses precision, not model validity or
study sample size. Every simulation estimate has Monte Carlo
uncertainty. Simulation-study guidance recommends reporting this
uncertainty and choosing the number of repetitions to achieve a stated
precision [@siepe_simulation_2024].

The **`plot_stability()`** checks if the estimated rejection probability
stabilizes as simulations accumulate. Compare the final value to the
target power (0.80). By default, it diagnoses each comparator separately
(`overall = FALSE`), use find which individual comparator or endpoint is
causing instability. On the other hand use `overall = TRUE` for
study-level precision checks, i.e to show the probability that the
complete all-comparators decision passes in a trial.

```{r workflow-stability, eval=TRUE, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_stability(ss_pilot, overall = TRUE)
```

The **`plot_mc_error()`**: Displays the remaining Monte Carlo
uncertainty around the power estimate. The title and labels above the
final points show the achieved Monte Carlo error (half-width of the 95%
confidence interval)

```{r workflow-mc-error, eval=TRUE, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_mc_error(ss_pilot, overall = TRUE)
```

Here, you need to ensure the curve has broadly stabilized and inspect
the displayed Monte Carlo error. If the error is too large for the
desired precision, increase `nsim`. For example, an estimate of 0.80
[0.79, 0.81] (half-width \~0.01) meets a 1% precision criterion, while
0.80 [0.765, 0.835] (half-width \~0.035) does not. In the latter case,
increase `nsim`—not the participant sample size.

**Note:** Small fluctuations in the curves are normal. Focus on the
final confidence interval width and broad stabilization. If power is
below the target power despite precision, increase the participant
sample size. If checks fail, correct model inputs and rerun.

## Step 5: Check Type I error at the equivalence boundaries

This step answers: "If the true estimand (in this example ratio of means
(ROM)) is exactly at an equivalence boundary, how often does the study
incorrectly conclude equivalence?"

### One endpoint example

Let us start with a small example so the calculation is easy to see.
Consider a trial with:

- **One comparator** (`T_vs_R1`) and **one endpoint** (`y1`).

- **Estimand, ROM**: $\theta = \frac{\mu_{T,y1}}{\mu_{R1,y1}}.$

- **Equivalence limits**: Lower (`L = 0.80`) and upper (`U = 1.25`).

- **Equivalence condition**: $L = 0.80, \qquad U = 1.25.$

The **TOST (Two One-Sided Tests) decision** requires rejecting both:

$$H_{0L}: \theta \leq L \quad\text{versus}\quad H_{1L}: \theta > L,$$
$$H_{0U}: \theta \geq U \quad\text{versus}\quad H_{1U}: \theta < U.$$

For a **log-normal outcome** with `ctype = "ROM"`, the boundary is set
on the **log-analysis scale**. Define

$$\eta_a = \log(\mu_{a,y1}) - \frac{1}{2}
\log\left(1 + \frac{\sigma_{a,y1}^2}{\mu_{a,y1}^2}\right),$$

where $\eta_a$ is the log-analysis mean for arm $a$.

For **lower-boundary scenario**, the function sets

$$\exp(\eta_T - \eta_{R1}) = L = 0.80.$$

With $\mu_{R1,y1} = 1.05$ and both standard deviations equal to 0.20,
this corresponds approximately to

$$\mu_{T,y1}=0.8478135,
\qquad
\frac{\mu_{T,y1}}{\mu_{R1,y1}}=0.8074414,
\qquad
\exp(\eta_T-\eta_{R1})=0.80.$$

For the **upper-boundary scenario**, the function sets:
$$\exp(\eta_T - \eta_{R1}) = U = 1.25.$$

For the same supplied reference mean and standard deviation, this gives
approximately $\mu_{T,y1}=1.304387$ and an arithmetic mean ratio of
1.242273, while the log-analysis ROM is exactly 1.25.

A **false-equivalence event** occurs if the confidence interval for the
estimated log-ratio falls within `(log(0.80), log(1.25))` despite the
true ROM being `0.80` or `1.25`.

Particularly, the `type1Error()` function simulates each scenario
individually, reporting the **type I error,** i.e. the
probability that all required tests incorrectly conclude equivalence
when the true effect is at the boundary of the acceptable range. For
example, if a study requires three endpoints to pass equivalence tests,
this error measures how often all three tests pass incorrectly at the
boundary.

When applied to the simss object with `null = "both"`, the function
evaluates both the lower and upper boundaries separately. With
`joint = TRUE`, it assesses every comparator–endpoint combination and
reports the **Type I error** for the complete trial decision, using the
stored k values.

The output table and plot display for each scenario the type I error along with **95% Monte
Carlo confidence intervals.**

```{r workflow-type1-one-scenario-plot, eval=TRUE, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
ss_one <- sampleSize(
  power = 0.80, alpha = 0.05,
  mu_list = list(T = mu_t["y1"], R1 = mu_r1["y1"]),
  sigma_list = list(T = sigma_t["y1"], R1 = sigma_r1["y1"]),
  list_comparator = list(T_vs_R1 = c("T", "R1")),
  list_y_comparator = list(T_vs_R1 = "y1"),
  list_lequi.tol = list(T_vs_R1 = c(y1 = 0.80)),
  list_uequi.tol = list(T_vs_R1 = c(y1 = 1.25)),
  dtype = "parallel", distribution = "lnorm", ctype = "ROM",
  lower = 10, upper = 100, nsim = 1000, seed = 2026, ncores = 1
)
type1_one <- type1Error(x = ss_one, null = "both", joint = TRUE)
print(type1_one)
plot(type1_one)
```

To estimate the **global Type I error**, use the least favorable
scenario (the one with the highest type I error). In the plot,
this scenario is highlighted with a thicker point and confidence line.
The target quantity is

$$\alpha_{\mathrm{global}}
 = \sup_{\theta \in H_0}
 P_{\theta}(\text{complete-trial success}),$$

the highest probability of a false complete-trial conclusion over the
null parameter space. This is the standard definition of the size of a
test for a composite null hypothesis [@lehmann2005, Chapter 3]. In this
multiple-endpoint setting, the least-favorable-configuration literature
applies the same principle by searching for the null configuration that
maximizes the rejection probability [@ristl2019]. In the one-endpoint
example, the boundary scenarios are the complete null configuration.

The `type1Error()` function evaluates the prespecified boundary grid and
reports the **largest empirical complete-trial success probability** as an
approximation to the **global Type I error**; probabilities are therefore not
averaged across scenarios. This is a finite-grid estimate, not a proof that
the reported scenario is the supremum over all effect sizes and nuisance
parameters.

Finally, compare the **global Type I error** point estimate and
confidence interval against the nominal value (here 5%):

- **Contains the nominal value**: Compatible with the nominal value at this
  simulation precision; this is not a formal calibration test.

- **Entirely above the nominal value**: Evidence of possible global inflation,
  subject to Monte Carlo error and the adequacy of the scenario grid.

- **Entirely below the nominal value**: Evidence of conservatism, subject to
  Monte Carlo error and the adequacy of the scenario grid.

### Multiple endpoints example

The one-endpoint example is illustrative, but in more complex cases
(e.g., the pilot example), the analysis extends to **every endpoint in
both comparisons** and for **both boundaries** (lower and upper).

For multiple endpoints, the null configuration depends on the decision
rule: when all endpoints are required (`k = m`), the global null is the
union of the component nulls; when only `k < m` endpoints are required,
it is a partial-conjunction null.

Ristl et al. [@ristl2019] describe the least-favorable-configuration
principle for multiple endpoints. In the all-required case, a candidate
least-favorable configuration has one component at its null boundary and
the other components at favorable alternatives.
For a `k`-of-`m` rule, the corresponding minimal null has
$$r = m - k + 1 $$ non-equivalent endpoints at boundaries and the remaining
`k - 1` endpoints at favorable alternatives.

For one fixed selected comparator, the number of configurations is
$$ {m \choose r}2^{r}. $$
Here, ${m \choose r}$ chooses the `r` boundary endpoints and $2^r$ chooses
whether each boundary is lower (`L`) or upper (`U`). If the analysis has $q$
comparators and the grid repeats these configurations with each comparator
selected in turn, the total number of configurations is
$$ q {m \choose r}2^{r}. $$
For example, with two comparators and three endpoints, this gives 12, 24,
and 16 configurations for `k = 3`, `k = 2`, and `k = 1`, respectively.
Additional nuisance-parameter values may be added for robustness. The
configuration grid is a finite approximation to the null parameter space.

Particularly, for each scenario, the **selected component** is placed at
its equivalence boundary, while
**other components** are set at favorable alternatives, here the
midpoint of their equivalence interval:

- For **continuous outcomes**: `(L + U) / 2` (difference) or
  `sqrt(L * U)` (ratio).

- For **log-normal ROM**, the midpoint is converted to an arithmetic
  mean using variance-based calibration.

- For **Poisson and negative-binomial outcomes**, `sqrt(L * U)` is used,
  but absolute rates, exposure, and dispersion (for negative binomial)
  affect precision.

The continuous-outcome midpoint is motivated by Anderson's inequality:
for a fixed-covariance Gaussian estimator, the probability of falling in
a symmetric convex acceptance region is maximized when its mean is at
the centre of that region [@anderson1955]. This gives `(L + U) / 2` on a
difference scale and `sqrt(L * U)` on a ratio scale, because the latter
is the midpoint on the log-ratio scale. In a multiple-endpoint
equivalence problem, this is used as a reproducible candidate
configuration for the non-selected components, alongside the
joint-equivalence framework described by Quan et al. [@quan2001]. It is
not a universal theorem for Poisson or negative-binomial models: their
nuisance rates, exposure, and dispersion can change the joint
probability, so those configurations may need an additional grid or
optimization check.

The midpoint is a practical finite candidate for a favorable
alternative, not a universal least-favorable result for Poisson or
negative-binomial models. For count outcomes, the grid should be
expanded over plausible rates, exposure, dispersion, and dependence
values when those nuisance parameters could change the maximizing
configuration.


### `k = 3`: all endpoints are required

When `k = m`, the decision for the complete trial is an
**intersection--union test (IUT)**. For three endpoints and two
comparisons, the trial passes only when

$$D = (D_{R1,y1} \cap D_{R1,y2} \cap D_{R1,y3})
    \cap
    (D_{R2,y1} \cap D_{R2,y2} \cap D_{R2,y3})$$

is true. The global null is the union of the component nulls: at least
one of the six component equivalence claims is false. Since the final
decision requires an intersection of component decisions, testing each
component TOST at level α does not require an additional endpoint-wise
multiplicity correction in this all-required setting. This is the usual
IUT logic for multiple-endpoint equivalence testing [@berger1996].

For `k = 3`, one endpoint is placed at a boundary. For `T_vs_R1`, the
six configurations are

``` text
(L, M, M)  (U, M, M)
(M, L, M)  (M, U, M)
(M, M, L)  (M, M, U)
```

For each of these, the `T_vs_R2` endpoints are `(M, M, M)`. The same six
configurations are then repeated with `T_vs_R2` as the selected null
comparator and `T_vs_R1` equal to `(M, M, M)`.

```{r workflow-type1-allscenario-plot, eval=TRUE, fig.width=8, fig.height=10, out.width="100%", fig.align="center"}
type1_global <- type1Error(x = ss_pilot, null = "both", joint = TRUE)
print(type1_global)
plot(type1_global)
```

The corresponding simulated error is the probability that the complete
trial decision passes, not merely the probability that the boundary
component passes.

### `k = 2`: at least two endpoints are required

The claim is that at least two of three endpoints are equivalent. This
is a **partial-conjunction** rule, not the all-required IUT. Its null
requires at least $r = 3 - 2 + 1 = 2$ non-equivalent endpoints.
Therefore, for `T_vs_R1`, the 12 candidate configurations are

``` text
(L, L, M)  (L, U, M)  (U, L, M)  (U, U, M)
(L, M, L)  (L, M, U)  (U, M, L)  (U, M, U)
(M, L, L)  (M, L, U)  (M, U, L)  (M, U, U)
```

In each case, `T_vs_R2` is `(M, M, M)`. Repeating the 12 configurations
with `T_vs_R2` selected gives

$$2\times {3\choose2}\times 2^2=24\ \text{scenarios}.$$

Because the decision is “at least two,” a false claim can occur when an
interior endpoint passes and one of the two boundary endpoints passes
falsely. Therefore, an endpoint-wise multiplicity adjustment is needed
for `k < m`; `adjust = "bon"` is a conservative starting point, and the
complete-trial Type I error must be checked empirically
[@benjamini2008].

```{r workflow-type1-allscenario-k2-plot, eval=TRUE, fig.width=8, fig.height=10, out.width="100%", fig.align="center"}
type1_global_k2 <- type1Error(x = ss_pilot_k2, null = "both", joint = TRUE)
print(type1_global_k2)
plot(type1_global_k2)
```

The default joint plot displays only the minimal null configurations, with
`m - k + 1` boundary endpoints (two endpoints here). The numerical global
Type I error still uses the complete grid, including configurations with
three boundary endpoints. To display every evaluated configuration, use
`plot(type1_global_k2, null_count = "all")`.

### `k = 1`: at least one endpoint is required

The claim is that at least one endpoint is equivalent. Its null requires
all three endpoints to be non-equivalent, so $r = 3 - 1 + 1 = 3$. For
`T_vs_R1`, the eight candidate configurations are

``` text
(L, L, L)  (L, L, U)  (L, U, L)  (L, U, U)
(U, L, L)  (U, L, U)  (U, U, L)  (U, U, U)
```

In each case, `T_vs_R2` is `(M, M, M)`. Repeating these eight
configurations with `T_vs_R2` selected gives

$$2\times {3\choose3}\times 2^3=16\ \text{scenarios}.$$

```{r workflow-type1-allscenario-k1-plot, eval=TRUE, fig.width=8, fig.height=10, out.width="100%", fig.align="center"}
type1_global_k1 <- type1Error(x = ss_pilot_k1, null = "both", joint = TRUE)
print(type1_global_k1)
plot(type1_global_k1)
```

This is the most permissive rule: any one boundary endpoint passing
falsely can complete the claim. A multiplicity adjustment is therefore
essential, and the global result should be checked using the largest
joint Type I-error estimate and its Monte Carlo interval.

For every configuration above, the simulated trial must satisfy the
complete study decision: the required number of endpoints must pass for
`T_vs_R1` **and** for `T_vs_R2` in the same trial.

The reported global Type I error is an **empirical approximation** of
the worst-case scenario among the evaluated boundary configurations.
Compare the global interval to the nominal value
(**5%) line** as we did in the previous section.

For `k = m`, `adjust = "no"` is generally
appropriate for the endpoint IUT. For `k < m`, use Mielke's strong
`adjust = "t"` calibration `alpha / (m - k + 1)` when strong control over
partial null configurations is required. The package's `adjust = "k"` method
uses `k * alpha / m` and is a weak complete-null calibration; it should be
used only when that estimand is the prespecified objective. Confirm the global
performance with `type1Error()`.

## Step 6: Run the final sample-size calculation

Only after the pilot has passed the distribution, estimand, correlation,
Monte Carlo, and Type I-error checks should the model be used for final
planning. The final call can use a larger `nsim` than the pilot.
Retaining raw data is optional at this stage; it is useful if the final
result also needs an audit trail of simulated outcomes.

```{r workflow-sample-size, eval=TRUE}
ss <- sampleSize(
  power = 0.80, alpha = 0.05,
  mu_list = list(T = mu_t, R1 = mu_r1, R2 = mu_r2),
  sigma_list = list(T = sigma_t, R1 = sigma_r1, R2 = sigma_r2),
  rho = 0.50, list_comparator = comparators,
  list_y_comparator = list(T_vs_R1 = endpoints, T_vs_R2 = endpoints),
  list_lequi.tol = lower, list_uequi.tol = upper,
  dtype = "parallel", ctype = "ROM", distribution = "lnorm",
  adjust = "no", k = 3, ncores = 1, nsim = 2000, seed = 2026,
  keep_sim_data = TRUE
)
```

In this example, the final run uses more trials than the 500-trial
pilot. If the final power decision is close to 0.80, increase this to
5,000 trials and use the final value displayed by `plot_mc_error(ss)` as
a stricter precision check. The appropriate choice is the smallest
`nsim` that gives a stable estimate and a confidence interval narrow
enough for the planned conclusion.

## Step 7: Interpret and report the final result

Now inspect the proposed sample size, achieved power, and Monte Carlo
interval. For a parallel design, `n` is the base sample size used to
derive the arm allocations; the summary also reports the total sample
size.

```{r workflow-result, eval=TRUE, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
print(ss)
summary(ss)
confint(ss)
plot(ss)
```

The result is acceptable only if the model checks passed, the Monte
Carlo interval is sufficiently narrow, the Type I-error assessment is
reasonable, and achieved power is close to the target. If the final
power is too low or too imprecise, increase the planned sample size or
`nsim` as appropriate. If it differs because a diagnostic failed,
correct the model inputs and rerun the full workflow rather than simply
increasing the sample size.

Finally, inspect endpoint-level decisions to identify which endpoint or
comparison drives failures of the global rule. This is a diagnostic for
the final result, not a replacement for the joint power or Type I-error
assessment.

In addition, the **`plot_decision_heatmap()`** helps identify which
endpoints or comparisons cause failures in the global decision rule by
visualizing pass (green) or fail (orange) results for each simulated
trial. Each row represents an endpoint, each facet a comparator (e.g.,
`T_vs_R1`), and the **Total row** shows the combined decision for that
comparator—green only if all required endpoints pass. If a specific
endpoint or comparator has more orange cells, it may be the bottleneck.
Use this alongside `summary(ss)` and `plot(ss)`: the summary provides
numerical power, while the heatmap highlights which endpoints or
comparisons are failing. Isolated orange cells are normal, but
persistent bands indicate systematic issues.

```{r workflow-heatmap, eval=TRUE, fig.width=8, fig.height=5, out.width="100%", fig.align="center"}
plot_decision_heatmap(ss, display = c("T_vs_R1", "T_vs_R2"))
```

# References
